# *Pseudomonas putida* grown in two soil types at two temperatures {#sec-26C-food-fuel-prac2}

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("drafting")
```

## Scenario

*Pseudomonas putida*, is a hydrocarbon-degrading bacterium used for
cleaning up oil-contaminated soil. *P.putida* was growth on either
hexadecane contaminated or control soil at two temperatures: 4 degrees
or 25 degrees. The was is to determine the effect of temperature and
soil type on the number of bacteria in the soil.

Each group in a large class did one replicate and added their data to a
Google sheet:
<https://docs.google.com/spreadsheets/d/1k2Gq4tKGEI0v_tdfgI-1d-Lk4YaWAC50sHffOmnqESU/edit?gid=886771226#gid=886771226>

```{r}
library(googlesheets4)
library(tidyverse)
library(emmeans)
```

```{r}
file <- "https://docs.google.com/spreadsheets/d/1k2Gq4tKGEI0v_tdfgI-1d-Lk4YaWAC50sHffOmnqESU/edit?gid=886771226#gid=886771226"
```

```{r}
class_data <- read_sheet(file) |> janitor::clean_names()
```

You will be asked to authenticate in your browser. This message will
appear in the console and the browser should open

```         
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
```

You can Allow “Tidyverse API Packages wants to access your Google
Account”.

You might also be asked

```         
Is it OK to cache OAuth access credentials in the folder
```

Choose the option for yes

```         
The httpuv package enables a nicer Google auth experience, in many cases, but it
isn't installed.
Would you like to install it now?
```

Choose the option for yes.


The data should then read in.

::: callout-note

## Next time you run `read_sheet()` from `googlesheets4`

Next time you run `read_sheet()` from `googlesheets4` you will
not have to authenticate in your browser if you chose "yes" for
`Is it OK to cache OAuth access credentials in the folder`.

Instead you will see:
```
The googlesheets4 package is requesting access to your Google account.
Enter '1' to start a new auth process or select a pre-authorized account.
1: Send me to the browser for a new auth process.
2: your.email@whereever.ac.uk
```
You can chose 2

:::

We don't need the timestamp or group name columns and these can be 
dropped using the `select()` command

The data are in wide format - there is sample in column. We want to
convert to long format so that each row contains a single count
and the other columns give the soil and the temperature treatment.
We can use `pivot_longer()` for this.

Drooping the columns (`-`) and pivoting the data can be done in one
pipeline:

```{r}
class_data <- class_data |> 
  select(-timestamp, -group_name) |>
  pivot_longer(cols = everything(),
               names_to = "treatment",
               values_to = "counts_cpg") 
```

```{r}

class_data <- class_data |>
  extract(
    col = treatment,
    into = c("soil", "temp"),
    regex = "^([uh]).([fi])_cells_gram_of_soil$"
  ) |>
  mutate(
    soil = recode(soil,
                  "u" = "uncontaminated",
                  "h" = "hexadecane contaminated"),
    temp = recode(temp,
                         "f" = "4 degrees",
                         "i" = "25 degrees")
  ) 

```

```{r}
ggplot(class_data, 
       aes(x = soil, y = counts_cpg, fill = temp)) +
  geom_boxplot() 

```

The data are not normally distributed - there are a few very large
values. Logging the data will aid visualisation:

```{r}
ggplot(class_data, 
       aes(x = soil, y = log10(counts_cpg), fill = temp)) +
  geom_boxplot() 

```

That's better but the warning:
`Removed 11 rows containing non-finite  outside the scale range`
indicates that there are some zeros. We can add 1 to the counts before
logging to avoid this. This is a common practice. It is useful because
`log(1) = 0` which simplifies interpretation and is better than removing
them. Many of the values are very large - the next smallest number is
75 - and adding 1 to large values before logging will have very little
impact.

```{r}
ggplot(class_data, 
       aes(x = soil, y = log10(counts_cpg + 1), fill = temp)) +
  geom_boxplot() 

```

Temperature seems to matter but there is little difference between the
two soils. The difference between the two temperatures is about the same
in both soils suggesting no interaction.

Add a new column to the data frame with the logged values that we can
use in the model.

```{r}
class_data <- class_data |> 
  mutate(log_counts = log10(counts_cpg + 1))
```

Summarise the data to get the means and se for each group.

```{r}
class_data_summary <- class_data |> 
  group_by(soil, temp) |> 
  summarise(mean = mean(log_counts),
            sd = sd(log_counts),
            n = length(log_counts),
            se = sd/sqrt(n)) 
```

two-way anova

```{r}
mod <- lm(data = class_data,
          log_counts ~ soil * temp)
```

```{r}
summary(mod)
```

```{r}
anova(mod)
```

There is a strong and significant effect of temperature, no effect of
soil and no interaction between temperature and soil. The effect of
temperature is independent of soil.

examine the assumptions

```{r}
ggplot(mapping = aes(x = mod$residuals)) +
  geom_histogram(bins = 30) 
```

Too may values in the middle, too few in the shoulders for a normal
distribution.

```{r}
shapiro.test(mod$residuals)
```

Significantly different from a normal distribution.

```{r}
plot(mod, which = 1)
```

Variances are similar in each group

Not normal but reasonably symmetrical and variances are similar in each
group. In addition, the effect is very clear - not at all borderline -
so this case I will continue with the two-way anova.

Post-hoc tests. Expecting differences between temps regardless of soil
type.

```{r}
emmeans(mod, ~ soil * temp) |> pairs()
```

```{r}

ggplot() + 
  geom_point(data = class_data, 
             aes(x = soil, y = log_counts, shape = temp), 
             position = position_jitterdodge(dodge.width = 1,
                                             jitter.width = 0.3, 
                                             jitter.height = 0), 
             size = 3, 
             colour = "gray50") +
  geom_errorbar(data = class_data_summary, 
                aes(x = soil, ymin = mean - se,
                    ymax = mean + se, group = temp), 
                width = 0.5, 
                linewidth = 1, 
                position = position_dodge(width = 1)) + 
  geom_errorbar(data = class_data_summary,
                aes(x = soil, ymin = mean, ymax = mean, group = temp), 
                width = 0.4, 
                linewidth = 1, 
                position = position_dodge(width = 1) ) +
  scale_x_discrete(name = "soil", expand = c(0, 0)) +
  scale_y_continuous(name = "Log Counts", expand = c(0, 0),                                                                                                              limits = c(0, 14)) +
  scale_shape_manual(values = c(19, 1), 
                     name = "Temperature") +
  # hexadecane contaminated 25 degrees - 
  # hexadecane contaminated 4 degrees <.0001 
  annotate("segment", 
           x = 0.75, xend = 1.25,
           y = 11.7, yend = 11.7, 
           colour = "black") +
  annotate("text", x = 1, y = 12, 
           label = "p < 0.0001") +
  # uncontaminated 25 degrees - 
  # uncontaminated 4 degrees <.0001 
  annotate("segment", 
           x = 1.75, xend = 2.25, 
           y = 11.7, yend = 11.7, 
           colour = "black") +
  annotate("text", x = 2, y = 12, 
           label = "p < 0.0001") +
  # hexadecane contaminated 25 degrees -
  # uncontaminated 4 degrees <.0001 
  annotate("segment", 
           x = 0.75, xend = 2.25, 
           y = 12.7, yend = 12.7, 
           colour = "black") +
  annotate("text", x = 1.5, y = 13, 
           label = "p < 0.0001") +
  # uncontaminated 25 degrees - 
  # hexadecane contaminated 4 degrees <.0001 
  annotate("segment", 
           x = 1.25, xend = 1.75, 
           y = 10.7, yend = 10.7, 
           colour = "black") +
  annotate("text",
           x = 1.5, y = 11, 
           label = "p < 0.0001") +
  theme_classic() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.92, 0.9),
        legend.background = element_rect(colour = "black"))

```

