# Association: Correlation and Contingency

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("drafting")
```

## Overview

no explanatory variables

### Correlation

-   two continuous variables

-   neither is an explanatory variable, i.e., there is not a causal relationship
    between the two variables
    
-   you want to know if there is a correlation between the two variables

-   the correlation coefficient is a measure of the strength and direction of
    the linear association between two variables and ranges from -1 to 1
    
-   A correlation coefficient of 1 indicates a perfect positive association 
    between the variables meaning the highest score in one variable are 
    associated with the *highest* scores in the other.

-   A correlation coefficient of -1 indicates a perfect negative association 
    between the variables meaning the highest scores in one variable are 
    associated with the *lowest* scores in the other.
    
TODO Add a figure of the different correlation coefficients
    
-   we estimate the population correlation coefficient, $\rho$ (pronounced rho), 
    with the sample correlation coefficient, $r$ and test whether it 
    is significantly different from zero because a correlation coefficient of
    0 means association.

-   Pearson's correlation coefficient, is a parametric measure which 
    assumes the variables are normally distributed and that any correlation
    is linear.

-   Spearman's rank correlation coefficient, is a non-parametric measure
    which does not assume the variables are normally distributed
    
-   we use `cor.test()` in R.

### Contingency Chi-squared



### Reporting


1.  the significance of effect - whether the association is significant
    different from zero

2.  the direction of effect
    
    -   correlation whether $r$ is positive or negative
    -   which categories have higher than expected values

3.  the magnitude of effect

We do not put a line of best fit on a scatter plot accompanying a correlation 
because such a line implies a causal relationship. 



We will explore all of these ideas with an examples.

## 🎬 Your turn!

If you want to code along you will need to start a new [RStudio
project](workflow_rstudio.html#rstudio-projects), add a `data-raw`
folder and open a new script. You will also need to load the
**`tidyverse`** package [@tidyverse].

## Correlation

High quality images of the internal structure of wheat seeds were taken with a 
soft X-ray technique. Seven measurements determined from the images:

-   Area
-   Perimeter
-   Compactness
-   Kernel length
-   Kernel width
-   Asymmetry coefficient
-   Length of kernel groove

Research were interested in the correlation between compactness and
kernel width

The data are in [seeds-dataset.xlsx](data-raw/seeds-dataset.xlsx).

### Import and explore

The data are in an excel file so we will need the **`readxl`** [@readxl] package.

Load the package:

```{r}
library(readxl)
```

Find out the names of the sheets in the excel file:


```{r}
excel_sheets("data-raw/seeds-dataset.xlsx")
```
There is only one sheet called `seeds_dataset`. 

Import the data:

```{r}
seeds <- read_excel("data-raw/seeds-dataset.xlsx",
                    sheet = "seeds_dataset")
```


Note that we could have omitted the `sheet` argument because there is only one 
sheet in the Excel file. However, it is good practice to be explicit. 

```{r}
#| echo: false

knitr::kable(seeds) |> 
  kableExtra::kable_styling() |> 
  kableExtra::scroll_box(height = "200px")
```

These data are formatted usefully for analysis using correlation and plotting.
Each row is a different seed and the columns are the variables we are interested. 
This means that the first values in each column are related - they are all from 
the same seed

In the first instance, it is sensible to create a rough plot of our data
(See @fig-seeds-rough). Plotting data early helps us in multiple ways:

-   it helps identify whether there extreme values
-   it allows us to see if the association is roughly linear
-   it tells us whether any association positive or negative

Scatter plots (`geom_point()`) are a good choice for exploratory
plotting with data like these.

```{r}
#| label: fig-seeds-rough
#| fig-cap: "A default scatter plot of compactness and kernel width in seeds demonstrates an approximately linear positive association them the variables."

ggplot(data = seeds,
       aes(x = compactness, y = kernel_width)) +
  geom_point()
```

The figure suggests that there is a positive correlation between `compactness` and
`kernel_width` and that correlation is roughly linear.


### Check assumptions

We can see from @fig-seeds-rough that the relationship between `compactness` and
`kernel_width` is roughly linear. This is a good sign for using Pearson's
correlation coefficient.

Our next check is to use common sense: both variables are continuous and
we would expect them to be normally distributed. We can then plot histograms
to examine the distributions (See @fig-hist).

```{r}
#| label: fig-hist
#| layout-ncol: 2
#| fig-cap: "The distributions of the two variables are slightly skewed but do not seem too different from a normal distribution. We will continue with the Pearson's correlation coefficient."
#| fig-subcap: 
#|   - "Compactness"
#|   - "Kernel width"

ggplot(data = seeds, aes(x = compactness)) +
  geom_histogram(bins = 12)

ggplot(data = seeds, aes(x = kernel_width)) +
  geom_histogram(bins = 12)

```
The distributions of the two variables are slightly skewed but do not seem too 
different from a normal distribution. 

Finally, we can use the Shapiro-Wilk test to test for normality.

```{r}
shapiro.test(seeds$compactness)
```

```{r}
shapiro.test(seeds$kernel_width)

```


The *p*-values are greater than 0.05 so these tests of the normality
assumption are not significant. Note that "not significant" means not
significantly different from a normal distribution. It does not mean
definitely normally distributed.

It is not unreasonable to continue with the Pearson's correlation coefficient. 
However, later we will also use the Spearman's rank correlation coefficient, a non-parametric method which has fewer assumptions. Spearman's rank correlation 
is a more conservative approach.


### Do a Pearson's correlation test with `cor.test()`

We can carry out a Pearson's correlation test with `cor.test()`  like this:

```{r}
cor.test(data = seeds, ~ compactness + kernel_width)
```

A variable is not being explained in this case so we do not need to include
a response variable. Pearson’s is the default test so we do not need to specify
it.

What do all these results mean?

The last line gives the correlation coefficient, $r$, that has been estimated 
from the data. Here, $r$ = `r cor.test(data = seeds, ~ compactness + kernel_width)$estimate |> round(2)` 
which is a moderate positive correlation.

The fifth line tells you what test has been done: 
`alternative hypothesis: true correlation is not equal to 0`. That is $H_0$ is 
that $\rho = 0$ 

The forth line gives the test result:
`t = 7.3738, df = 68, p-value = 2.998e-10`

The $p$-value is very much less than 0.05 so we can reject the null hypothesis 
and conclude there is a significant positive correlation between `compactness` 
and `kernel_width`.

### Report

There was a significant positive correlation ($r$ = `r cor.test(data = seeds, ~ compactness + kernel_width)$estimate |> round(2)`) between compactness and kernel width (Pearson's: *t* = 7.374; *df* = 68; *p*-value < 0.0001). See @fig-correlation-pears.


::: {#fig-correlation-pears}
```{r}
#| code-fold: true

ggplot(data = seeds,
       aes(x = compactness, y = kernel_width)) +
  geom_point() +
  scale_y_continuous(name = "Diameter (mm)") +
    scale_x_continuous(name = "Compactness") +
  annotate("text", x = 0.85,  y = 3.6, 
           label = expression(italic(r)~"= 0.67; "~italic(p)~"< 0.001")) +
  theme_classic()
```


**Correlation between compactness and kernel width of wheat seeds**. High
quality images of the internal structure of wheat seeds were taken with a 
soft X-ray technique and the compactness and kernel width of the seeds were
determined. There was a significant positive correlation ($r$ = `r cor.test(data = seeds, ~ compactness + kernel_width)$estimate |> round(2)`) between compactness 
and kernel width (Pearson's: *t* = 7.374; *df* = 68; *p*-value < 0.0001). Note:
axes do not start at 0. Data analysis was conducted in R [@R-core] with
tidyverse packages [@tidyverse].

:::

### Spearman's rank correlation coefficient

## Contingency Chi-squared test

