# Ideas about data {#sec-ideas-about-data}

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("complete")
```

## Introduction

This chapter introduces some key ideas about data. Data consists of:

-   Variables — the properties or characteristics that we measure or
    record, and
-   Observations — the individual cases or items that those variables
    describe.

A helpful and common way to organise data is in a table, where each
column represents a variable and each row represents one observation.
Every value in a row related to a single observation such as a person,
plant, or sample. We call data in this format "tidy".

We can describe variables in two main ways:

-   By their role in the analysis, that is, whether they are independent
    (explanatory) or dependent (response) variables.

-   By the type of values they take and how often those values occur.

The second aspect — what kinds of values a variable can have and how
common each one is — is called the variable’s distribution. The way we
summarise, visualise, and analyse a variable depends on both its role
and its distribution.

## Roles of variables in analysis

When we do research, we typically have variables that we choose or set
and variables that we measure. The variables we choose or set are called
independent or explanatory variables. The variables we measure are
called dependent or response variables. For example, we might measure
the concentration of enzymes and hormones in blood samples of
individuals with different genotypes. The genotype acts as the
explanatory variable and the blood measurements are response. We would
be interested in whether the blood measures differ between genotypes.

## Distributions

Any variable has a *distribution* which describes the values the
variable can take and the chance of them occurring. The distribution is
a function (relationship) which captures how the values are mapped to
the probability of them occurring. A distribution has a general type,
given by the function and is further tuned by the *parameters* in the
function. For example, variables like height, length, concentration,
mass and intensity follow a *normal distribution* also know as the
bell-shaped curve so that they look similar. However, they have
different means and standard deviations. The mean and the standard
deviation are the parameters of the normal distribution.

An important distinction is between **discrete** and **continuous**
types of data. Continuous variables are measurements that can take *any*
value in their range. Because there an infinite number of possible
values within within the range, the probability of a continuous variable
taking exactly one specific value is zero. Instead, we talk about the
probability that the value is less than a particular value, greater than
a particular value or between two values. For example, we might say the
probability a person's height is greater than 170cm is 0.3
(@fig-distrib-contin). In contrast, discrete variables can take only
specific values often counts or categories. For example, the number of
leaves on a plant or the number of times a gene is expressed. In this
case, it's meaningful to talk about the probability of a single value.
Each possible value has its own individual probability, and the sum of
all those probabilities equals 1 (@fig-distrib-discrete).

::: {#fig-distrib-contin}
```{r}
#| echo: false

ggplot(data = data.frame(x = c(-3, 3)),
       aes(x)) +
  stat_function(fun = dnorm_limit,
                args = list(q = -0.7, tails = 1),
                geom = "area",
                fill = pal2[1]) +
  stat_function(fun = dnorm, n = 101) +
  annotate("text", x = -1.7, y = 0.3, 
           label = "Shaded area under curve\nis the probability of getting\na value of X or less",
           size = 4) +
  annotate("segment", xend = -1.1, yend = 0.1, 
           x = -1.7, y = 0.25, 
           linewidth = 1,
           colour = pal4[1],
           arrow = arrow(type = "open", 
                         length = unit(0.05, "npc"))) +
  scale_y_continuous(breaks = NULL, name = "",
                     expand = c(0, 0),
                     limits = c(0, 0.5)) +
  scale_x_continuous(breaks = c(-0.7), 
                     labels = c("X"),
                     name = "Value a continuous variable can take",
                     expand = c(0, 0),
                     limits = c(-3, 3)) +
  theme_classic() +
  theme(axis.line.y = element_blank(),
        axis.ticks = element_blank())

```

**An example of a distribution for a continuous variable.** The values a
continuous variable can take are on the horizontal axis. Continuous
variables have a smooth distribution and the probability of getting a
particular value or less is given by the area under the curve before
that value. This example is the Normal distribution.
:::

::: {#fig-distrib-discrete}
```{r}
#| echo: false

mean <- 2
ggplot(data.frame(x = c(0:6)), aes(factor(x), fill = factor(x))) +
  geom_col(aes(y = dpois(x, mean)), colour = "black" ) +
  annotate("text", size = 4,
           x = 4.5, y = 0.26,
           label = "Probability of getting\na value of exactly Y") +
  annotate("segment",
           x = 4.4, y = 0.23,
           xend = 4, yend = 0.15,
           linewidth = 1,
           colour = pal4[1],
           arrow = arrow(type = "closed", 
                         length = unit(0.03, "npc"))) +
  scale_y_continuous(breaks = NULL, 
                     name = "",
                     expand = c(0, 0),
                     limits = c(0, 0.3)) +
  scale_x_discrete(name = "Value a discrete variable can take",
                  breaks = 3, labels = "Y") +
  scale_fill_manual(values = c("white",
                               "white",
                               "white",
                               pal2[1],
                               "white",
                               "white",
                               "white")) +
  theme_classic() +
  theme(legend.position = "none")

```

**An example of a distribution for a discrete variable.** A discrete
variable has a stepped distribution and the probability of getting
exactly a particular is given by the area of the bar at that value. This
example is the Poisson distribution.
:::

This difference between continuous and discrete variables affects how we
analyse and visualise data. Discrete data is often shown using bar plots
and summarised with modes of medians, while continuous data is
represented using histograms or density plots and summarised with means
and standard deviations.

## Summarising data

We summarise data by giving a measure of *central tendency* and a
measure of *dispersion*. A measure of central tendency is a single value
that represents the middle or centre of a variable's distribution. The
three most common measures are:

-   the mean, or more accurately, the arithmetic mean,
    $\bar{x} = \frac{\sum{x}}{n}$

-   the median: the middle value for a variable with values arranged in
    order of magnitude

-   the mode: the most frequent value in the variable.

Measures of dispersion describe the spread of values around the centre
and indicate how much variability there is in the variable. The most
common measures of dispersion are:

-   the range: the difference between the maximum value and the minimum
    value in a variable

-   the interquartile range: two values, the first quartile and the
    third quartile. The first quartile is half way between the median
    value and the lowest value when the values are arranged in order and
    the third quartile is halfway between the median value and the
    highest value

-   the variance: the average of the squared differences between each
    value and the variable's mean,
    $s^2 = \frac{(\sum{x - \bar{x})^2}}{n - 1}$

-   the standard deviation: the square root of the variance.

## Discrete data

Discrete variables can take only specific values and can be categories,
like genotype, or counts, like the number of petals.

### Categories: Nominal and Ordinal

Categorical data can be nominal or ordinal depending on whether the
categories are are ordered. Nominal variables have no particular order,
for example, the eye colour of Drosophila or the species of bird. When
summarising data on bird species, it wouldn't matter in what order the
information was given or plotted. Ordinal variables have an order. The
Likert scale [@likert1932] used in questionnaires is one example. The
possible responses are Strongly agree, Agree, Disagree and Strongly
disagree; these have an order that you would use when plotting them
(@fig-categories).


::: {#fig-categories layout-nrow="1"}

```{r}
#| echo: false
#| label: fig-categ-bird
birds <- data.frame(Response = c("Robin",
                                 "Chaffinch",
                                 "Thrush",
                                 "Wren"),
                    Frequency = c(12, 6, 10, 13))

birds |> 
  ggplot(aes(x = Response, y = Frequency)) +
  geom_col(fill = pal2[1]) +
  scale_x_discrete(name = "Species") +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0 , 15)) +
  theme_classic(base_size = 16) 
```

```{r}
#| echo: false
#| label: fig-categ-likert
likert <- data.frame(Response = factor(c("Strongly agree",
                                         "Agree",
                                         "Disagree",
                                         "Strongly disagree"), 
                                       levels = c("Strongly agree",
                                                  "Agree",
                                                  "Disagree",
                                                  "Strongly disagree")),
                     Frequency = c(10, 14, 6, 2))

likert |> 
  ggplot(aes(x = Response, y = Frequency)) +
  geom_col(fill = pal2[2]) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0 , 15)) +
 theme_classic(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, 
                                   vjust = 1, 
                                   hjust = 1)) 


```

**Categorical data can be nominal or ordinal.** Nominal variables like 
species (@fig-categ-bird) have no particular order whilst ordinal variables
like those on the Likert scale [@likert1932] have an order.

:::

The most appropriate way to summarise nominal or ordinal data is to
report the mode (most frequent value) or tabulate the number of each
value (@tbl-likert).

```{r}
#| echo: false
#| label: tbl-likert
#| tbl-cap: "Frequency of responses to a Likert scale question."

knitr::kable(likert) |> 
  kableExtra::kable_styling()

```

### Counts

Counts are one of the most common data types. They are quantitative but
discrete because they can take only specific values. Counts tend to
follow a distribution called the Poisson distribution meaning that there
is an expected number of zeros, ones and twos etc that appear. This is
true no matter what you count. The distribution of counts is not
symmetrical when the average count is low. For example, the number of 
groundsel plants in one hundred 1 metre x 1 metre square 
quadrats[^ideas_about_data-1] is likely in the range of 0 to 3 plants 
per quadrat. This distribution is skewed to the right, meaning it has a
tail to the right. A mean is not usually the best way to summarise a 
set of counts because it is dragged up by the few quadrats with more 
than 2 plants (@fig-counts).  Instead, the median better represents the 
middle of the distribution. The median is the middle value when the 
values are arranged in order. The interquartile range (IQR) indicates 
the spread in the data. The lower quartile is half way between the 
lowest value and the middle value and the upper quartile is half way 
between the middle value and the highest value.


[^ideas_about_data-1]: A quadrat is a frame used in ecology to outline a
    standard unit of area for study of the distribution of plant species
    over a wider are. Typically, quadrats are placed randomly over the
    area and the number of individuals in each quadrat is recorded.


::: {#fig-counts}

```{r}
#| echo: false

set.seed(5)
nquad <- 100
mu <- 0.7
plants <- data.frame(nplants = rpois(nquad, mu)) 
sample_m <- mean(plants$nplants)
sample_median <- median(plants$nplants)

plants |> ggplot(aes(x = factor(nplants))) +
  geom_bar(fill = pal2[3]) +
  annotate("segment", x = sample_m + 1, xend = sample_m + 1, 
           y = 0, yend = 58,
           colour = pal4[1],
           linewidth = 1) +
  annotate("segment", x = sample_median + 1, xend = sample_median + 1, 
           y = sample_median, yend = 58,
           colour = pal4[2],
           linewidth = 1) +
  annotate("text", x = sample_m + 1, y = 60,
           label = paste("mean = ", sample_m),
           colour = pal4[1]) +
  annotate("text", x = 1, y = 60,
           label = paste("median = ", sample_median),
           colour = pal4[2]) +
  scale_x_discrete(name = "Number of plants per quadrat") +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0 , 65),
                     name = "Number of quadrats") +
  theme_classic() +
  theme(legend.position = "none")

```

**The number of groundsel plants in one hundred 1 metre x 1 metre square 
quadrats is a count and follows a Poisson distribution.** The mean of 
`r sample_m` groundsel plants per quadrat does not really reflect that the 
majority of quadrats contained no plants! The average is being dragged up by 
the few quadrats containing 3 plants. The median is `r sample_median`.

:::



## Continuous data

Continuous variables are measurements that can take *any* value in their
range so there are an infinite number of possible values. The values
have decimal places. Variables like the length and mass of an organism,
the volume and optical density of a solution, or the colour intensity of
an image are continuous. Many response variables are continuous but
continuous variables can also be explanatory. A large number of
continuous variables follow the normal distribution.

### The normal distribution

For example, a variable like human height has values with decimal places
which follow a normal distribution also known as the Gaussian
distribution or the bell-shaped curve. Values of 1.65 metres occur more
often than values of 2 metres and values of 3 metres never occur
(@fig-normal-distribtion).

::: {#fig-normal-distribtion}

```{r}
#| echo: false
m <- 1.65
sd <- 0.06
ggplot(data = data.frame(Height = c(m - 3 * sd, m + 3 * sd)), aes(Height)) +
  stat_function(fun = dnorm, n = 101, 
                args = list(mean = m, sd = sd)) +
  scale_y_continuous(breaks = NULL, name = "Likelihood of value occuring",
                     expand = c(0, 0)) +
  annotate("text", x = 1.5, y = 4.5,
           label = "Values are rare") +
    annotate("text", x = 1.65, y = 4.5,
           label = "Values are common") +
    annotate("text", x = 1.8, y = 4.5,
           label = "Values are rare") +
  theme_classic()
```

**Human height follows a normal distribution.**

:::

The mean and standard deviation are the best way to summarise a normally
distributed variable.



## Theory and practice {#sec-ideas-about-data-theory-practice}

The distinction between continuous and discrete values is clear in
theory but in practice, the actual values you have may differ from the
expected distribution for a particular variable. For example, we would
expect the mass of cats to be continuous because any value in the range
is possible. However, if our scales only measure to the nearest
kilogram, then the values become discrete because the only the values
possible are 0, 1, 2, 3, 4, 5, 6 and 7. The gap between values, 1kg, is
big compared to the range of values we expect (@fig-contin-is-discrete).

::: {#fig-contin-is-discrete}

```{r}
#| echo: false

m <- 4
sd <- 0.8
set.seed(1234)
a <- ggplot(data = data.frame(Mass = c(m - 3 * sd, m + 3 * sd)), aes(Mass)) +
  stat_function(fun = dnorm, n = 101, 
                args = list(mean = m, sd = sd)) + 
  scale_y_continuous(breaks = NULL, name = "", 
                     expand = c(0, 0)) +
  annotate("text", x = m - 2 * sd, y = 0.4,
           label = "Theory") +
  theme_classic()

b <- ggplot(data = data.frame(Mass = round(rnorm(1000, m, sd), 0)), aes(Mass)) +
  geom_histogram(binwidth = 1, colour = "black", fill = "white") +
  scale_y_continuous(breaks = NULL, name = "",
                     expand = c(0, 0)) +
  annotate("text", x = m - 2.5 * sd, y = 400,
           label = "Practice when\nmeasured to\nthe nearest 1kg") +
  theme_classic()

a + b
```


**The mass of cats in kilograms.** Theoretically mass is a continuous variable
(left). However if we measure only to the nearest kilogram, then mass 
becomes discrete because cats only weigh few kilograms so 1kg covers a 
large chunk of the range (right).
:::



In contrast, the number of hairs on a human head ranges from about
80,000 to 150,000 so the difference between having 100,203 hairs and
100,204 hairs is so tiny the variable is practically continuous
(@fig-discrete-is-contin).

::: {#fig-discrete-is-contin}

```{r}
#| echo: false

m <- 120000
sd <- 20000
set.seed(12)
ggplot() +
  geom_histogram(data = data.frame(hairs = round(rnorm(60000, m, sd), 0)),
                 aes(hairs),
                 bins = 120, colour = "black", fill = "white") +
  scale_y_continuous(breaks = NULL, name = "",
                     expand = c(0, 0)) +
  scale_x_continuous("Number of hairs on head") +
  theme_classic()


```

**The number of hairs on a human head**. As a count, this is a discrete 
variable. However. one increment is a tiny proportion of the range 
of values so the number of hairs is practically continuous.

:::

A rule of thumb is that if the mean count is above about 100, then the
distribution of counts closely matches the normal distribution.

## Summary

1.  Data: Data consists of variables (properties measured or recorded)
    and observations (individual things with those properties). Data is
    commonly organised with variables in columns and observations in
    rows.

2.  Roles of Variables in Analysis: Variables in research can be
    independent or explanatory variables (chosen or set by researchers)
    and dependent or response variables (measured by researchers).

3.  Distributions: A variable's distribution describes the types of
    values it can take and the likelihood of each value occurring.
    Variables can be classified as discrete (specific values) or
    continuous (any value within a range). The shape of a distribution
    is determined by parameters.

4.  Discrete Data: Discrete variables can be categories or counts.
    Categorical data can be nominal (no order) or ordinal (ordered) and
    are best summarised with the mode or a table. Counts are often best
    summarised by the median and interquartile range.

5.  Continuous Data: Continuous variables can take any value within a
    range. The normal distribution is the most common continuous
    distribution and is best summarised with the mean and standard
    deviation.

6.  Theory and Practice: While the theoretical distinction between
    continuous and discrete values is clear, the actual values obtained
    in practice may deviate from the expected distribution. Measurement
    precision and range affect whether a variable is considered
    continuous or discrete. A rule of thumb suggests that if the mean
    count is above approximately 100, the distribution of counts
    approximates a normal distribution.
