# About this book

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("drafting")
```

## Who is this book for?

bioscience

undergrads

## About the content
It is in sections

part 1 what they forgot to teach you

focus on what causes problems for people learning to code.

part 2 Getting started with data. give summary

part 3 Data Analysis, improve name, give summary (babs 2)

## Your turn!
When you see.. Try to answer before looking at the code

🎬 Your turn! Assign the value of `4` to a variable called `y`:

```{r}
#| code-fold: true
y <- 4
```



🎬 Your turn! Use the file system above to answer these questions.

-   What is the absolute path for the document` doc4.txt` on a Mac computer?

::: {.callout-tip collapse="true"}

## 📖

-  `/home/user1/docs/data/doc4.txt` 

:::






conventions used in this book


## Commenting
This page has commenting with Hypothes.is enabled 

make a private group
https://web.hypothes.is/help/annotating-with-groups/

## Code of Conduct for commenting