# About this book

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("drafting")
```

## Who is this book for?

bioscience

undergrads


## Approach of this book


##  Overview of the contents

It is in sections

part 1 what they forgot to teach you

focus on what causes problems for people learning to code.

part 2 Getting started with data. give summary

part 3 Data Analysis, improve name, give summary (babs 2)




## Conventions used in the book

Code and any output appears in blocks formatted like this:

```{r}
#| eval: false

stag <- read_table("data-raw/stag.txt")
glimpse(stag)
```

Lines of output start with a `#`. 

Within the text:
-  packages are indicated in bold code font like this: **`ggplot2`**
-  functions are indicated in code font with brackets after their name like this: `ggplot()`
-  R objects are indicated in code font like this: `stag`


The content of a code block can be copied using the icon in its top right corner.

I use packages from the **`tidyverse`** [@tidyverse] including **`ggplot2`** [@ggplot2], **`dplyr`** [@dplyr], **`tidyr`** [@tidyr] and **`readr`** [@readr] throughout the book. All the code assumes you have loaded the core **`tidyverse`** packages with: 

```{r}
#| eval: false

library(tidyverse)
```

If you run examples and get an error like this: 

```{r}
#| eval: false

# Error in read_table("data-raw/stag.txt") : 
#  could not find function "read_table"
```

It is likely you need to load the **`tidyverse`** as shown above.

All other packages will be loaded explicitly with `library()` statements where needed. 



🎬 Your turn! indicates that you might want to code along with examples or that there is an opportunity to check your understanding by answering a question. Questions are answered in words or with a piece of code. The answers are given in collapsed sections so you can try to answer them before checking the answer. For example, a question answered in words looks like this:


🎬 Your turn! Use the file system above to answer these questions.

-   What is the absolute path for the document` doc4.txt` on a Mac computer?

::: {.callout-tip collapse="true"}

## 📖

-  `/home/user1/docs/data/doc4.txt` 

:::

And a question answered with a piece of code looks like this:


🎬 Your turn! Assign the value of `4` to a variable called `y`:

```{r}
#| code-fold: true
#| 
y <- 4
```





## Commenting
This page has commenting with Hypothes.is enabled 

make a private group
https://web.hypothes.is/help/annotating-with-groups/

## Code of Conduct for commenting

## Credits

This book is written with R [@R-core], Quarto [@allaire2022], `knitr` [@knitr], `kableExtra` [@kableExtra]. My R session information is shown below:

```{r}
sessionInfo()
```
