# About this book {#sec-intro}

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("polishing")
```

## Who is this book for?

This book is primarily to support [Bioscience
students](https://www.york.ac.uk/biology/) at the [University of
York](https://www.york.ac.uk/). The ultimate aim is to support the full
spectrum of computational skills that a bioscience undergraduate or
postgraduate at York - and elsewhere - might need. This is live book with 
ideas being added and material drafted -- and hopefully improved -- over time. 
Each pages is labelled with its status as one of the following:

```{r}
#| results: "asis"
#| echo: false
status("drafting")
status("polishing")
status("complete")
```



The content included so far is described in the [Overview of
contents](#overview-of-contents) section below.



It is being written in the open so that it can be used by anyone who
finds it useful. It is also being written in the open so that anyone can
contribute to it.

## Approach of this book

-   explanations followed by worked examples

## Overview of contents {#overview-of-contents}

It is in sections

**Section 1: What they forgot to teach you about computers**

This chapter tries to teach the computer skills that you might have
missed if you have used mainly the mobile devices. I focus on the
knowledge gaps that often appear when people are learning computational
data analysis. Primarily these are to do with finding and organising
their files and folders in the file systems.

**Section 2 Getting started with data analysis in R**

The first steps into analysing data with R. The first chapter in this
part covers important concepts about data: whether they are discrete and
continuous and how we summarise them using descriptive statistics. These
ideas are not specific to analysing data with R, they apply whatever
analysis tool you use. The second chapter introduces you to R and
RStudio for the first time. It explains what R, RStudio and R packages
are and where to find them. It then explores the layout and appearance.
In the third chapter you will start coding and lean the most common data
types and data structures used for data analysis. The fourth chapter
describes some useful workflow patterns and tools for organising your
work in RStudio. Using these will make learning R easier. Finally, we
will go through a complete workflow from importing data from a file to
saving a figure for reporting.

**Section 3 Statistical Analysis in R - Part 1**

This section is a first course in Statistical inference which is the
process of inferring the characteristics of populations from samples
using data analysis. In this first course we take what is called a
frequentist - or classical - approach to statistical inference. This is
the approach that is most commonly taught in introductory statistics
courses. We will learn about the logic of hypothesis testing and
confidence intervals. You will also get an introduction to statistical
models, what is a statistical model and in particular a linear model.

**Section 4 Statistical Analysis in R - Part 2**



**Section 5 Working with Gel data in R**

## Conventions used in the book

I use some conventions most of which I hope are intuitive. I have tried
to articulate them here. If you recognise conventions I have used that
are not listed here please [let me know](#contributing).

Code and any output appears in blocks formatted like this:

```{r}
# import the chaff data
chaff <- read_table("data-raw/chaff.txt")
glimpse(chaff)
```

Lines of *output* start with a `##` to distinguish from code *comments*
which begin with a single `#`. You will learn more about comments in the
[Using Scripts](#using-scripts) section in [First Steps in
RStudio](#first_steps_rstudio.html)

Within the text:

-   packages are indicated in bold code font like this: **`ggplot2`**
-   functions are indicated in code font with brackets after their name
    like this: `ggplot()`
-   R objects are indicated in code font like this: `stag`

The content of a code block can be copied using the icon in its top
right corner.

I use packages from the **`tidyverse`** [@tidyverse] including
**`ggplot2`** [@ggplot2], **`dplyr`** [@dplyr], **`tidyr`** [@tidyr] and
**`readr`** [@readr] throughout the book. All the code assumes you have
loaded the core **`tidyverse`** packages with:

```{r}
#| eval: false
library(tidyverse)
```

If you run examples and get an error like this:

```{r}
#| eval: false

# Error in read_table("data-raw/stag.txt") : 
#  could not find function "read_table"
```

It is likely you need to load the **`tidyverse`** as shown above.

All other packages will be loaded explicitly with `library()` statements
where needed.

When you see "🎬 Your turn!" indicates that you might want to code along
with examples or that there is an opportunity to check your
understanding by answering a question. Questions are answered in words
or with a piece of code. The answers are given in collapsed sections so
you can try to answer them before checking the answer. For example, a
question answered in words looks like this:

🎬 Your turn! Use the file system above to answer these questions.

-   What is the absolute path for the document`doc4.txt` on a Mac
    computer?

::: {.callout-tip collapse="true"}
## 📖

-   `/home/user1/docs/data/doc4.txt`
:::

And a question answered with a piece of code looks like this:

🎬 Your turn! Assign the value of `4` to a variable called `y`:

```{r}
#| code-fold: true
#| 
y <- 4
```

## Annotating this book

This page has annotating with [Hypothesis](https://web.hypothes.is/)
enabled. Hypothesis allows you to annotate this book with your own
private notes or make notes shared with friends. You need to create a
free personal account. You can make annotations that are public, private
only to you or shared with a [private
group](https://web.hypothes.is/help/annotating-with-groups/). Please
follow the code of conduct in your annotations.

## Code of Conduct

We are dedicated to providing a welcoming and supportive learning
environment for all readers, regardless of background or identity. As
such, we do not tolerate comments that are disrespectful to fellow
learners or that excludes, intimidates, or causes discomfort to others.
The following bullet points set out explicitly what we hope you will
consider to be appropriate community guidelines:

-   Be respectful of different viewpoints and experiences. Do not use in
    homophobic, racist, transphobic, ageist, ableist, sexist, or
    otherwise exclusionary language.

-   Use welcoming and inclusive language. Do not address others in an
    angry, intimidating, or demeaning manner. Be considerate of the ways
    the words you choose may impact others. Be patient and respectful of
    the fact that English is a second (or third or fourth!) language for
    many.

-   Respect the privacy and safety of others. Do not share their
    information without their express permission.

-   As an overriding general rule, please be intentional in your actions
    and humble in your mistakes.

## Contributing {#contributing}

This book is being written in the open so that anyone can contribute to
it. If you find a mistake, or have a suggestion for improvement you can
[create an issue](https://github.com/3mmaRand/comp4biosci/issues/new).

## License

<p xmlns:cc="http://creativecommons.org/ns#">

This work is licensed under
<a href="http://creativecommons.org/licenses/by-nc/4.0/?ref=chooser-v1" target="_blank" rel="license noopener noreferrer" style="display:inline-block;">CC
BY-NC
4.0<img src="https://mirrors.creativecommons.org/presskit/icons/cc.svg?ref=chooser-v1" style="height:22px!important;margin-left:3px;vertical-align:text-bottom;"/><img src="https://mirrors.creativecommons.org/presskit/icons/by.svg?ref=chooser-v1" style="height:22px!important;margin-left:3px;vertical-align:text-bottom;"/><img src="https://mirrors.creativecommons.org/presskit/icons/nc.svg?ref=chooser-v1" style="height:22px!important;margin-left:3px;vertical-align:text-bottom;"/></a>
This license requires that reusers give credit to the creator. It allows
reusers to distribute, remix, adapt, and build upon the material in any
medium or format, for noncommercial purposes only.

</p>

## Please cite as

Please cite this book as:

Rand, E. (2025). Computational Analysis for Bioscientists (Version 0.2)
https://3mmarand.github.io/comp4biosci/

## Credits

This book is written with R [@R-core], Quarto [@allaire2022],
**`knitr`** [@knitr1; @knitr2; @knitr3], **`kableExtra`** [@kableExtra].
My R session information is shown below:

```{r}
sessionInfo()
```
