# Finding the molecular weight of proteins migrated on a western blot

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("drafting")
```

## What is a Western Blot?

A Western blot is a laboratory technique used to detect specific
proteins in a sample. Proteins are isolated from the sample, sorted
by size using SDS-PAGE and then transferred to a membrane (usually
nitrocellulose or PVDF). The membrane is incubated with a primary 
antibody that binds to the target protein. A secondary antibody, 
linked to a detectable enzyme or fluorophore, binds to the primary
antibody so the protein of interested can be detected. The result is a 
band on the membrane. The molecular weight of a protein can be estimated
from the position of the band on the membrane because the distance 
travelled by the protein is inversely proportional to the log of the 
molecular weight. The intensity of the band can also be used to estimate
the amount of protein present in the sample.





## Scenario

Macrophages are immune cells that play a key role in the body's defense
against pathogens. They can be activated by various stimuli, including
bacterial infection. In this experiment, we are interested in the
production of TNF-alpha, a pro-inflammatory cytokine, in response to
bacterial infection. We will use a Western blot to determine the

Macrophages are treated with one of three treatments: Media, LPS or 
NeonGreen fluorescent *E.coli* Macrophages produce TNF-α in response to
bacterial infection

![Model western blot data containing protein weight markers (lane 1) and
anti-TNFalpha stained samples for media (in lane 2), LPS-stimulated (in
lane 3), and E.coli-stimulated (in lane 4)
samples.](data-images/western-blot-anti-TNFalpha%20stained-macrophages.png){width="200"}

![A picture of the protein weight markers used in the experiments with
their associated weights in kilodaltons
(kDa).](data-images/protein-weight-markers.png){width="200"}

```{r}
library(tidyverse)
library(imager)
```

Use the **`imager`** package to import your Western blot image.

```{r}
blot <- load.image("data-images/western-blot-anti-TNFalpha stained-macrophages.png")
```


```{r}
plot(blot)
```

For a typical blot, y increases downwards, so smaller bands (higher kDa) are higher up.


Annotate the Lanes and Marker Lane

```{r}
#| eval: false
# Number of bands in your marker lane
# click from the top or gel to the bottom 
# i.e., high MW to low
marker_positions <- locator(n = 11)  
```


```{r}
#| eval: false
# make sure these are in the same order 
# that you clicked lanes in
# i.e., high MW to low

marker_df <- tibble(
  x = marker_positions$x,
  y = marker_positions$y,
  # Replace with your actual marker weights
  KDa = c(250, 180, 130, 95, 72, 55, 43, 34, 26, 17, 10)  
)
```


```{r}
#| eval: false
migration_points <- locator(n = 2)
```


```{r}
#| eval: false

origin_y <- migration_points$y[1]
front_y <- migration_points$y[2]
total_migration <- abs(front_y - origin_y)
```



```{r}
#| eval: false
# band positions for each sample: LPS, E.coli
tnf_positions <- locator(n = 2)  
```


```{r}
#| eval: false
tnf_df <- tibble(
  x = tnf_positions$x,
  y = tnf_positions$y,
  distance_migrated = abs(tnf_positions$y - origin_y),
  Rf = distance_migrated / total_migration
)
```



```{r}
#| eval: false
marker_df <- marker_df  |> 
  mutate(
    distance_migrated = abs(y - origin_y),
    Rf = distance_migrated / total_migration,
    log_kda = log10(KDa)
  )
```


```{r}
#| eval: false
ggplot(data = marker_df, aes(x = Rf, y = log_kda)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)


```


```{r}
#| eval: false
model <- lm(log_kda ~ Rf, data = marker_df)
summary(model)
```

```{r}
#| eval: false
# Predict TNF band molecular weight
tnf_df <- tnf_df  |> 
  mutate(
    predicted_log_kda = predict(model, newdata = tnf_df),
    predicted_kda = 10^predicted_log_kda
  )

```


TNF-alpha has a molecular weight of approximately 17 kDa for the 
monomeric form and 52 kDa for the homotrimer. The 17 kDa monomer 
is often observed on SDS-PAGE under reducing conditions, while 
the trimer is the form typically secreted. The predicted molecular 
weight of the precursor membrane-bound form is around 26 kDa. 

Thus, we have the precursor membrane-bound form.

```{r}
#| eval: false
ggplot(data = marker_df, aes(x = Rf, y = log_kda)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE,
              colour = "black",
              linetype = 2) +
   geom_point(data = tnf_df, 
              aes(x = Rf, y = predicted_log_kda), 
              shape = 23, size = 3) +
  theme_classic()


```


```{r}
#| eval: false
plot(blot, axes = FALSE)
points(marker_df$x, marker_df$y, pch = "--")
text(20, 
     marker_df$y, labels = marker_df$KDa, cex = 0.8)
text(c(140, 210, 275), 
     c(10,10, 10), labels = c("Media"), cex = 0.8)
```




