# Molecular weight from SDS-PAGE {#mol-wt-sds-page}

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("polishing")
```

## Overview

Sodium Dodecyl Sulfate-Polyacrylamide Gel Electrophoresis (SDS-PAGE) is
a method to separate proteins by size.

Proteins are denatured and coated with SDS (a detergent), giving them a
uniform negative charge. They migrate through a polyacrylamide gel when
an electric current is applied. The lower the molecular weight of the
protein, the faster, and therefore further, it migrates through the gel.
Gels are run with a "marker lane" containing proteins of known molecular
weights. The positions of these marker proteins are used to create a
"standard curve" that allows us to estimate the molecular weights of
other proteins in the gel. The proteins are made visible by staining
(e.g. with Coomassie Blue). SDS-PAGE can tell you the size of proteins
in a sample, but not their identity. It can also be used to find the
relative abundances of the proteins.

this section we will use a linear regression to estimate the molecular
weights of proteins from an SDS-PAGE gel.

This is a different use for linear regression than the single linear
regression we covered in @sec-single-linear-regression. In that section
we wanted to statistically test whether an *x*-variable explained the
variation in a *y*-variable.

Here we already know there is a very tight linear relationship between
the position of the marker proteins on the gel and their molecular
weights. But here we are using the linear regression to find the
equation of the line that describes this relationship so we can use it
to estimate the molecular weights of other proteins in the gel.

We will explore all of these ideas with an example.

## 🎬 Your turn!

If you want to code along you will need to start a new RStudio Project
(see [@sec-workflow-rstudio-creating-rsp]), add a `data-image` folder
and open a new script. You will also need to load the **`tidyverse`**
package [@tidyverse].

## Linear Regression for estimating molecular weights from gels

We will use this gel image (@fig-gel) as an example:
[sds-page-gel.jpg](data-image/sds-page-gel.jpg). This gel has two sets
of results (one each side) - since only 4 lanes were needed for each
experiment, we put two experiments on one gel.

-   Lane 1 (and 10) is the protein ladder which are the proteins of
    known molecular weights.

-   Lane 2 (and 9) is Uninduced E. coli lysate

-   Lane 3 (and 8) is Induced E. coli lysate

-   Lane 4 (and 7) is ShPatB

You don't need worry about the details of the gel, only that our aim is
to estimate the molecule weight of the ShPatB protein in lane 4 from its
position on the gel and the standard curve created from the marker
proteins in lane 1. @fig-gel illustrates the measurements needed from
the gel image.

```{r}
#| echo: false
#| message: false

# import gel
library(imager)
gel <- load.image("data-image/sds-page-gel.jpg")
gel_cropped <- crop.borders(gel, nx = 300, ny = 150)
gel_top <- 180
gel_bottom <- 990
pos_patB <- 394
```

::: {#fig-gel}
```{r}
#| echo: false
#| message: false

plot(gel_cropped, axes = FALSE)
# add a line to plot 
abline(h = gel_top, col = "red")
abline(h = gel_bottom, col = "red")
# draw vertical double-headed arrow between top and bottom
segments(x0 = 50, y0 = gel_top, 
       x1 = 50, y1 = gel_bottom, 
       lty = 2)
text(x = 20, y = (gel_top + gel_bottom) / 2, 
     labels = "L", 
     cex = 1.2)
arrows(x0 = 220, y0 = gel_top, 
       x1 = 220, y1 = gel_top + 138,
       col = "blue",
       code = 3, 
       length = 0.1)
arrows(x0 = 510, y0 = gel_top, 
       x1 = 510, y1 = gel_top + pos_patB,
       col = "blue",
       code = 3, 
       length = 0.1)
```

**SDS-PAGE gel provided as a sample in the BCH strand of 26C practicals
with annotation.** Lane 1 (and 10) is the protein ladder, Lane 2 (and 9)
is Uninduced E. coli lysate, Lane 3 (and 8) is Induced E. coli lysate
and Lane 4 (and 7) is ShPatB. The horizontal red lines indicate the top
and bottom of the gel. The vertical black dashed line indicates the
length of the gel, $L$. The distance of one marker protein - the
heaviest - and that ShPatB from the top of the gel are indicated by
vertical solid blue lines. Distance to all the marker proteins need to
be measured from the top of the gel. These distances and the molecular
weights of the marker proteins are used to create a standard curve that
allows us to estimate the molecular weight of ShPatB from its position
on the gel.
:::

We will cover two options:

1.  Where you have measured the length of the gel, the positions of
    ShPatB and the marker proteins on the gel manually - by opening the
    gel image in Powerpoint for example -- and have a file containing
    the molecular weights and positions of the marker proteins. The
    measures can be in centimetres or pixels, it does not matter as long
    as they are all in the same units.

2.  Where you have a file containing the molecular weights of the marker
    proteins and use R to measure the length of the gel, the positions
    of ShPatB and the marker proteins by importing the gel image. This
    method relies on the `locator()` function which stores the
    coordinate positions when you click on a plot! Magic!

### Option 1: Manual measurements

You have measured the positions of the marker proteins and added them to
a file containing the molecular weights. Your file is:
[standard-mw-with-positions.csv](data-raw/standard-mw-with-positions.csv).
The molecular weights of the marker proteins are in kilodaltons (kDa)
and the positions of the marker proteins are in pixels. The length of
the gel is 810 pixels and ShPatB is at 394 pixels from the top of the
gel.

Save
[standard-mw-with-positions.csv](data-raw/standard-mw-with-positions.csv)
to `data-raw/` and import it.

```{r}
mw_positions <- read_csv("data-raw/standard-mw-with-positions.csv")
```

Assign the position of ShPatB and the length of the gel to variables:

```{r}
gel_length <- 810
pos_patB <- 394
```

We need to calculate $R_f$ values for the marker proteins:

$$R_f = \frac{L - d}{L}$$

where $L$ is the length of the gel and $d$ is the distance to the band.

We also need to log the molecular weights of the marker proteins to make
a linear relationship. We can add these two new variables to the data
frame using the `mutate()`.

Calculate $R_f$ values for the marker proteins:

```{r}
mw_positions <- mw_positions |>
  mutate(Rf = (gel_length - dist_to_band) / gel_length,
         log_kda = log(kda))
```

Plot the data with `geom_point()` and add a linear regression line with
`geom_smooth(method = "lm")`:

```{r}
ggplot(mw_positions, aes(x = Rf, y = log_kda)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE) +
  theme_classic()
```

Fit a linear model so we have the equation of the line

```{r}
mod <- lm(log_kda ~ Rf, data = mw_positions)
```

Print the model:

```{r}
mod
```

We only need to print the coefficients - we don't care about the
statistical tests here. You can tell from the plot that the relationship
is very tight and linear. The equation of the line is: $MW$=
`r mod$coef[2] |> round(3)` $\times R_f$ + `r mod$coef[1] |> round(3)`

You can substitute the values of the coefficients and the $R_f$ of
ShpatB to find the log molecular weight of ShPatB. Or you can use the
`predict()` function to do this for you.

Calculate the $R_f$ of ShPatB:

```{r}
patB_Rf = (gel_length - pos_patB) / gel_length

```

Predict the molecular weight ShPatB:

```{r}
patB_kda <- predict(mod, newdata = data.frame(Rf = patB_Rf)) |> 
  exp()
patB_kda
```

### Option 2: Measurements in R

You have a file containing the molecular weights of the marker proteins
[standard-mw.txt](data-raw/standard-mw.txt) and the image of the gel
[sds-page-gel.jpg](data-image/sds-page-gel.jpg). The molecular weights
of the marker proteins are in kilodaltons (kDa)

Make a folder call `data-image` and save
[sds-page-gel.jpg](data-image/sds-page-gel.jpg) to it. Save
[standard-mw.txt](data-raw/standard-mw.txt) to `data-raw`.

Import the molecular weights of the marker proteins from
[standard-mw.txt](data-raw/standard-mw.txt)

```{r}
mw <- read_csv("data-raw/standard-mw.txt")
```

Load the `imager` package

```{r}
library(imager)
```

Import the gel image:

```{r}
gel <- load.image("data-image/sds-page-gel.jpg")
```

Base R's generic `plot()` function can handle image files and plot axes
default. These are in pixels and will help us mark the top and bottom of
the gel.

Plot the gel image:

```{r}
plot(gel)
```

The **`imager`** package has a function that will crop the edges of the
image. This is certainly not essential but can have two benefits:

-   the image is smaller which means plotting is quicker -- this is
    especially useful when your images are many pixels
-   it makes it a little easier determine where the axes numbers are on
    the gel

Crop the image:

```{r}
# crop
gel_cropped <- crop.borders(gel, nx = 300, ny = 150)
```

`crop.borders()` removes `ny` pixels from the top and the bottom and
`nx` pixels from each side. You may wish to adjust the numbers.

Plot the cropped image:

```{r}
plot(gel_cropped)
```

To make sure we measure distances from the same place we need to add
lines to mark the top and the bottom of the gel. Notice that the
*y*-axis is 0 at the top. I think the top is at about 180 and the bottom
is about 990. We will assign these values to variables because they will
be needed in calculations later. We will also need the gel length
(bottom position - top position).

Assign values for the top and bottom of the gel to variables and
calculate the length of the gel:

```{r}
gel_top <- 180
gel_bottom <- 990
gel_length <- gel_bottom - gel_top
```

Plot the cropped gel with lines marking the top and bottom of the gel:

```{r}
# plot gel
plot(gel_cropped)
# add a horizontal lines for the top and bottom of the gel 
abline(h = gel_top, col = "red")
abline(h = gel_bottom, col = "red")
```

Make sure you run all these commands. The base plotting system works a
little differently than `ggplot`. We do not use `+` but we have to make
sure we have recently run the `plot()` command before the `abline()`
(and other functions that modify plots) will work. You will get
`Error in int_abline(a = a, b = b, h = h, v = v, untf = untf, ...) :`
`plot.new has not been called yet` if you have not recently run the
`plot()` command.

At this point you have check that you are happy with the numbers used
for the top and bottom of the gel. Adjust and replot if needed.

I like to add a vertical line in the marker lane to help guide my later
measures.

Add a vertical line in the marker lane. Again, make sure you run all of
the plotting code:

```{r}
# plot gel
plot(gel_cropped)
# add a horizontal lines for the top and bottom of the gel
abline(h = gel_top, col = "red")
abline(h = gel_bottom, col = "red")
# add a vertical line in the marker lane as a guide to help with locator
abline(v = 220, col = "red")
```

We are now ready to measure the band positions using the `locator()`
function. We need to measure the position of shPatB and the all the
marker proteins. We will start with shPatB:

Run the `locator()` command and click on the shPatB in lane 4:

```{r}
#| eval: false
dist_to_patB <- locator(n = 1)
```

```{r}
#| echo: false
dist_to_patB <- 394 + gel_top
```

The (*x*,*y*) position of shPatB is now stored in an R object called
`pos_patB`.

The distance between the shPatB band and the top of the gel will be the
*y* value in `dist_to_patB` minus the distance to the top of the gel.

Calculate the distance from the top of the gel to shPatB:

```{r}
#| eval: false
pos_patB <- dist_to_patB$y - gel_top
```

```{r}
#| echo: false
pos_patB <- 394
```

We have 9 bands so pass the argument `n = 9` to the function. This means
you will need to click on the graph 9 times. Start at the top - the
heaviest band - and work your way down. You only need to click once on
each band. The R cursor will disappear until you have clicked 9 times.
Don't worry if you make a mistake, just click until the cursor is
returned and run the locator command again to start afresh.

Run the `locator()` command and click on the 9 bands in order from top
to bottom:

```{r}
#| eval: false
# Number of bands in your marker lane
# click from the top or gel to the bottom 
# i.e., high MW to low
marker_positions <- locator(n = 9)  
```

```{r}
#| echo: false
temp <- read_csv("data-raw/standard-mw-with-positions.csv")
marker_positions <- list(x = rep(220, 9), y = temp$dist_to_band + gel_top)
```

Magic! The (*x*,*y*) position of each band is now stored in an R object
called `marker_positions`.

We need to

-   combine the *y* positions with molecular weights in `mw`
-   calculate the distance to each band by subtracting `gel_top`
-   calculate $R_f$ values for the marker proteins using
    $R_f = \frac{L - d}{L}$ where $L$ is the length of the gel and $d$
    is the distance to the band.
-   log the molecular weights of the marker proteins to make a linear
    relationship.

We can put all these new columns in a dataframe called `mw_positions`
frame using the `mutate()`.

Create `mw_positions` from `mw` and the *y* positions in
`marker_positions`:

```{r}
mw_positions <- mw |>
  mutate(y = marker_positions$y,
         dist_to_band = y - gel_top,
         Rf = (gel_length - dist_to_band) / gel_length,
         log_kda = log(kda)) 
```

The process is now exactly the same as it was for Option 1.

Plot the data with `geom_point()` and add a linear regression line with
`geom_smooth(method = "lm")`:

```{r}
ggplot(mw_positions, aes(x = Rf, y = log_kda)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE) +
  theme_classic()
```

Fit a linear model so we have the equation of the line

```{r}
mod <- lm(log_kda ~ Rf, data = mw_positions)
```

Print the model:

```{r}
mod
```

We only need to print the coefficients - we don't care about the
statistical tests here. You can tell from the plot that the relationship
is very tight and linear. The equation of the line is: $MW$=
`r mod$coef[2] |> round(3)` $\times R_f$ + `r mod$coef[1] |> round(3)`

You can substitute the values of the coefficients and the $R_f$ of
ShpatB to find the log molecular weight of ShPatB. Or you can use the
`predict()` function to do this for you.

Calculate the $R_f$ of ShPatB:

```{r}
patB_Rf = (gel_length - pos_patB) / gel_length

```

Predict the molecular weight ShPatB:

```{r}
patB_kda <- predict(mod, newdata = data.frame(Rf = patB_Rf)) |> 
  exp()
patB_kda
```

If you would like to practice Option 2 again, you could repeat the the
steps using the set of results on the other side of the gel.

## Summary

TO DO
