# One-way ANOVA and Kruskal-Wallis {#sec-one-way-anova-and-kw}

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("complete")
```

## Overview

In the last chapter, we learnt how to use and interpret the general
linear model when the *x* variable was categorical with two groups. You
will now extend that to situations when there are more than two groups.
This is often known as the one-way ANOVA (**an**alysis **o**f
**var**iance). We will also learn about the Kruskal-Wallis test
[@kruskal1952], a non-parametric test that can be used when the
assumptions of the general linear model are not met.

We use `lm()` to carry out a one-way ANOVA. General linear models
applied with `lm()` are based on the normal distribution and known as
parametric tests because they use the parameters of the normal
distribution (the mean and standard deviation) to determine if an effect
is significant. Null hypotheses are about a mean or difference between
means. The assumptions need to be met for the *p*-values generated to be
accurate.

If the assumptions are not met, we can use the non-parametric equivalent
known as the Kruskal-Wallis test. Like other non-parametric tests, the
Kruskal-Wallis test :

-   is based on the ranks of values rather than the actual values
    themselves
-   has a null hypothesis about the mean rank rather than the mean
-   has fewer assumptions and can be used in more situations
-   tends to be less powerful than a parametric test when the
    assumptions are met

<!-- Why not do several two-sample tests? ANOVA terminology and concepts -->

The process of using `lm()` to conduct a one-way ANOVA is very similar
to using `lm()` for a two-sample *t*-test, but with an important
distinction. When we obtain a significant effect of our explanatory
variable, it only indicates that at least two group means differ—it does
not specify which ones. To determine where the differences lie, we need
a **post-hoc test**.

A post-hoc ("after this") test is performed after a significant ANOVA
result. There are several options for post-hoc tests, and we will use
Tukey's HSD (honestly significant difference) test [@tukey1949],
implemented in the **`emmeans`** [@emmeans] package.

Post-hoc tests adjust *p*-values to account for multiple comparisons. A
Type I error occurs when we incorrectly reject a true null hypothesis,
with a probability of 0.05. Conducting multiple comparisons increases
the likelihood of obtaining a significant result by chance. The post-hoc
test corrects for this increased risk, ensuring more reliable
conclusions.

### Model assumptions

The assumptions for a general linear model where the explanatory
variable has two or more groups, are the same as for two groups: the
residuals are normally distributed and have homogeneity of variance.

If we have a continuous response and a categorical explanatory variable
with three or more groups, we usually apply the general linear model
with `lm()` and *then* check the assumptions, however, we can sometimes
tell when a non-parametric test would be more appropriate before that:

-   Use common sense - the response should be continuous (or nearly
    continuous, see Ideas about data: Theory and
    practice (see @sec-ideas-about-data-theory-practice). Consider
    whether you would expect the response to be continuous
-   There should decimal places and few repeated values.

To examine the assumptions after fitting the linear model, we plot the
residuals and test them against the normal distribution in the same way
as we did for single linear regression.

### Reporting

In reporting the result of one-way ANOVA or Kruskal-Wallis test, we
include:

1.  the significance of the effect

    -   parametric (GLM): The *F*-statistic and *p*-value
    -   non-parametric (Kruskal-Wallis): The Chi-squared statistic and
        *p*-value

2.  the direction of effect - which mean/median is greater in a the
    pairwise commparison

    -   Post-hoc test

3.  the magnitude of effect - how big is the difference between the
    means/medians

    -   parametric: the means and standard errors for each group
    -   non-parametric: the medians for each group

Figures should reflect what you have said in the statements. Ideally
they should show both the raw data and the statistical model:

-   parametric: means and standard errors
-   non-parametric: boxplots with medians and interquartile range

We will explore all of these ideas with some examples.

## 🎬 Your turn!

If you want to code along you will need to start a new RStudio Project
(see [@sec-workflow-rstudio-creating-rsp]), add a `data-raw`
folder and open a new script. You will also need to load the
**`tidyverse`** package [@tidyverse].

## One-way ANOVA

Researchers wanted the determine the best growth medium for growing
bacterial cultures. They grew bacterial cultures on three different
media formulations and measured the diameter of the colonies. The three
formulations were:

-   Control - a generic medium as formulated by the manufacturer
-   sugar added - the generic medium with added sugar
-   sugar and amino acids added - the generic medium with added sugar
    and amino acids

The data are in [culture.csv](data-raw/culture.csv).

In this scenario, our null hypothesis, $H_0$, is that there is no
difference in colony diameter between the three groups or that group
membership has no effect on diameter. This is written as:
$H_0: \beta_1 = \beta_2 = 0$

This means that none of the group means differ significantly from each
other.

Another way of describing, $H_0$, is to say the variation caused by
media is no greater than the the random background variation. This is
expressed with the ***F*****-statistic**, a variance ratio that compares
the variance explained by the model to the variance within groups. If
$H_0$ is true, the variance between groups is no greater than the
variance within groups. That is, *F*, $\frac{Var(between)}{Var(within)}$
equals 1. This is written as: $H_0: F = 1$

Both versions of the null hypothesis describe the same idea: that 
here is **no real difference** between the groups.

-   **Using $\beta$ coefficients:** This version focuses on the 
    **linear model**. It says that the group variable has no 
    effect, meaning all groups have the same average value. 
    If the coefficients ($\beta_1$, $\beta_2$, etc.) are **zero**, 
    then the model isn’t capturing any meaningful differences between 
    groups.

-   **Using the F-statistic:** This version looks at **variance**. It 
    compares the variation **between groups** to the variation 
    **within groups**. If the F-value is **close to 1**, it means the 
    differences between groups are about the same as random variation 
    within groups, so there’s no real effect.

Both approaches test the same hypothesis—if we reject the null, 
we conclude that at least one group is different from the others.


### Import and explore

Import the data:

```{r}
culture <- read_csv("data-raw/culture.csv")
```

```{r}
#| echo: false
knitr::kable(culture) |> 
  kableExtra::kable_styling() |> 
  kableExtra::scroll_box(height = "200px")
```

The Response variable is colony diameters in millimetres and we would
expect it to be continuous. The Explanatory variable is type of media
and is categorical with 3 groups. It is known “one-way ANOVA” or
“one-factor ANOVA” because there is only one explanatory variable. It
would still be one-way ANOVA if we had 4, 20 or 100 different types of media.

These data are in tidy format [@Wickham2014-nl] - all the diameter
values are in one column with another column indicating the media. This
means they are well formatted for analysis and plotting.

In the first instance it is sensible to create a rough plot of our data.
This is to give us an overview and help identify if there are any issues
like missing or extreme values. It also gives us idea what we are
expecting from the analysis which will make it easier for us to identify
if we make some mistake in applying that analysis.

Violin plots (`geom_violin()`, see @fig-culture-rough), box plots
(`geom_boxplot()`) or scatter plots (`geom_point()`) all make good
choices for exploratory plotting and it does not matter which of these
you choose.

:::{#fig-culture-rough}

```{r}

ggplot(data = culture,
       aes(x = medium, y = diameter)) +
  geom_violin()
```

**The diameters of bacterial colonies when grown in one of three media**. 
A violin plot is an alternative to a boxplot for getting an overview of our data
before we begin analysis. Quick graphical overviews help us identify any 
issues such as missing or extreme values. It also tells us what to expect from 
the analysis. Adding both sugar and amino acids to the medium seems to
increase the colony diameter.

:::

R will order the groups alphabetically by default.

The figure suggests that adding sugar and amino acids to the medium
increases the diameter of the colonies.

Summarising the data for each medium is the next sensible step. The most
useful summary statistics are the means, standard deviations, sample
sizes and standard errors. I recommend the `group_by()` and
`summarise()` approach:

```{r}
culture_summary <- culture |> 
  group_by(medium) |> 
  summarise(mean = mean(diameter),
            std = sd(diameter),
            n = length(diameter),
            se = std/sqrt(n))

```


```{r}
#| echo: false
# extract values for ease of reporting
cont_m <- culture_summary |> 
  filter(medium == "control") |> 
  pull(mean) |> round(2)

cont_se <- culture_summary |> 
  filter(medium == "control") |> 
  pull(se) |> round(2)

cont_n <- culture_summary |> 
  filter(medium == "control") |> 
  pull(n) 

suga_m <- culture_summary |> 
  filter(medium == "sugar added") |> 
  pull(mean) |> round(2)

suga_se <- culture_summary |> 
  filter(medium == "sugar added") |> 
  pull(se) |> round(2)

suga_n <- culture_summary |> 
  filter(medium == "sugar added") |> 
  pull(n)

suaa_m <- culture_summary |> 
  filter(medium == "sugar and amino acids added") |> 
  pull(mean) |> round(2)

suaa_se <- culture_summary |> 
  filter(medium == "sugar and amino acids added") |> 
  pull(se) |> round(2)

suaa_n <- culture_summary |> 
  filter(medium == "sugar and amino acids added") |> 
  pull(n)

```

We have saved the results to `culture_summary` so that we can use the
means and standard errors in our plot later.

```{r}
culture_summary
```

### Apply `lm()`

We can create a one-way ANOVA model like this:

```{r}
mod <- lm(data = culture, diameter ~ medium)
```

And examine the model with:

```{r}
summary(mod)
```

The Estimates in the Coefficients table give:

-   `(Intercept)` known as $\beta_0$. The mean of the control group
    (@fig-one-way-anova-lm-model). Just as the intercept is the value of
    the *y* (the response) when the value of *x* (the explanatory) is
    zero in a simple linear regression, this is the value of `diameter`
    when the `medium` is at its first level. The order of the levels is
    alphabetical by default.

-   `mediumsugar added` known as $\beta_1$. This is what needs to be
    added to the mean of the control group to get the mean of the
    'medium sugar added' group (@fig-two-sample-lm-model). Just as the
    slope is amount of *y* that needs to be added for each unit of *x*
    in a simple linear regression, this is the amount of `diameter` that
    needs to be added when the `medium` goes from its first level to its
    second level (*i.e.*, one unit). The `mediumsugar added` estimate is
    positive so the the 'medium sugar added' group mean is higher than
    the control group mean

-   `mediumsugar and amino acids added` known as $\beta_2$ is what needs
    to be added to the mean of the control group to get the mean of the
    'medium sugar and amino acids added' group
    (@fig-two-sample-lm-model). Note that it is the amount added to the
    *intercept* (the control in this case). The
    `mediumsugar and amino acids added` estimate is positive so the the
    'medium sugar and amino acids added' group mean is higher than the
    control group mean

If we had more groups, we would have more estimates and all would be
compared to the control group mean.

The *p*-values on each line are tests of whether that coefficient is
different from zero.

-   `(Intercept)                     10.0700     0.2930  34.370  < 2e-16 ***`
    tells us that the control group mean is significantly different from
    zero. This is not a very interesting, it just means the control
    colonies have a diameter.
-   `mediumsugar added                 0.1700     0.4143   0.410  0.68483`
    tells us that the 'medium sugar added' group mean is not
    significantly different from the control group mean.
-   `mediumsugar and amino acids added   1.3310     0.4143   3.212  0.00339 **`
    tells us that the 'medium sugar and amino acids added' group mean
    *is* significantly different from the control group mean.

Note: none of this output tells us whether the medium sugar and amino
acids added' group mean *is* significantly different from the 'medium
sugar added' group mean. We need to do a post-hoc test for that.

The *F* value and *p*-value in the last line are a test of whether the
model as a whole explains a significant amount of variation in the
response variable.

:::{#fig-two-sample-lm-model}

```{r}
#| echo: false

ggplot() +
  geom_errorbar(data = culture_summary, 
                aes(x = medium, ymin = mean, ymax = mean), 
                colour = pal3[1], linewidth = 1,
                width = 1) +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), 
                     limits = c(9, 12),
                     name = "Diameter (mm)") +
  geom_segment(aes(x = 0.8,
                   xend = -Inf,
                   y = mod$coefficients[1] + 1,
                   yend = mod$coefficients[1]),
               colour = pal3[2]) +
  annotate("text",
           x = 0.9, 
           y = mod$coefficients[1] + 1.15,
           label = glue::glue("Intercept (β0) is mean\nof { culture_summary$medium[1] }: { mod$coefficients[1]|> round(2) }"), 
           colour = pal3[2],
           size = 3) +
  geom_segment(aes(x = 1.5,
                   xend = 1.5,
                   y = mod$coefficients[1],
                   yend = mod$coefficients[1] + mod$coefficients[2]),
               colour = pal3[3],
               arrow = arrow(length = unit(0.03, "npc"), 
                             ends = "both")) +
  geom_segment(aes(x = 1.5,
                   xend = 1.7,
                   yend = mod$coefficients[1] - 0.25,
                   y = mod$coefficients[1] + 0.1),
               colour = pal3[2]) +
  annotate("text",
           x = 2, 
           y = mod$coefficients[1] - 0.35,
           label = glue::glue("mediumsugar added (β1) is the difference\nbetween { culture_summary$medium[1] } mean and { culture_summary$medium[2] } mean: { mod$coefficients[2] |> round(2) }"), 
           colour = pal3[2],
           size = 3) +
  geom_segment(aes(x = 3,
                   xend = 3,
                   y = mod$coefficients[1],
                   yend = mod$coefficients[1] + mod$coefficients[3]),
               colour = pal3[3],
               arrow = arrow(length = unit(0.03, "npc"), 
                             ends = "both")) +
  geom_segment(aes(x = 3,
                   xend = 2.1,
                   yend = mod$coefficients[1] + mod$coefficients[3] - 0.5,
                   y = mod$coefficients[1] + mod$coefficients[3] - 0.3),
               colour = pal3[2]) +
  annotate("text",
           x = 2.1, 
           y = mod$coefficients[1] + mod$coefficients[3] - 0.75,
           label = glue::glue("mediumsugar and amino acids added (β2)\nis the difference between { culture_summary$medium[1] } mean\nand { culture_summary$medium[3] } mean: { mod$coefficients[3] |> round(2) }"), 
           colour = pal3[2],
           size = 3) +
   geom_segment(aes(x = 1.5,
                   xend = 3,
                   yend = mod$coefficients[1],
                   y = mod$coefficients[1]),
               colour = pal3[1],
              linetype = "dashed") + 
theme_classic()

```


**The a linear model fitted to the relationship between media and diameter**.
In an one-way ANOVA model with three groups, the model has three coefficients
(parameters) estimated from the data. The first estimate is the intercept 
which is the mean of the first group. The second estimate is the 'slope' 
to the second group which is what has to added to the intercept to get 
the second group mean. The third estimate is the 'slope' to the third group
which is what has to added to the intercept to get the third group mean. 
Note that y axis starts at 15 to create more space for the annotations.

:::

If you find this a bit overwhelming to read you can use the `anova()`
function on the model object to get a simpler output:

```{r}
anova(mod)
```


```{r}
#| echo: false
# extract values from mod using broom for ease of reporting
mod_tidy <- broom::tidy(anova(mod))

F_value <- mod_tidy |> 
  filter(term == "medium") |> 
  pull(statistic) |> round(2)
p_value <- mod_tidy |> 
  filter(term == "medium") |> 
  pull(p.value) |> round(4)
df_effect <- mod_tidy |> 
  filter(term == "medium") |> 
  pull(df) 
df_residuals <- mod_tidy |>  
  filter(term == "Residuals") |> 
  pull(df)

```

The ANOVA is significant but this only tells us that growth medium
matters, meaning at least two of the means differ. To find out which
means differ, we need a post-hoc test.

### Post-hoc testing

A post-hoc ("after this") test is done after a significant ANOVA test. If the 
ANOVA is *not* significant, no post-hoc is needed.

There are several possible post-hoc tests and we will be using Tukey's HSD
(honestly significant difference) test [@tukey1949] implemented in 
the **`emmeans`** [@emmeans] package.

We need to load the package:

```{r}
library(emmeans)
```

Then carry out the post-hoc test:

```{r}
emmeans(mod, ~ medium) |> pairs()

```

```{r}
#| echo: false
# extract pvalues from emmeans using broom for ease of reporting
ph_tidy <- emmeans(mod, ~ medium) |> pairs() |> broom::tidy()

p_cont_v_suga <- ph_tidy |> 
  filter(contrast == "control - sugar added") |> 
  pull(adj.p.value) |> round(4)
p_cont_v_suaa <- ph_tidy |> 
  filter(contrast == "control - sugar and amino acids added") |> 
  pull(adj.p.value) |> round(4)
p_suga_v_suaa <- ph_tidy |> 
  filter(contrast == "sugar added - sugar and amino acids added") |> 
  pull(adj.p.value) |> round(4)


```


Each row is a comparison between the two means in the 'contrast' column.
The 'estimate' column is the difference between those means and the
'p.value' indicates whether that difference is significant. In this case,
two of the comparisons are significant: 

-  control - sugar and amino acids added with *p* = `r p_cont_v_suaa`
-  sugar added - sugar and amino acids added with *p* = `r p_suga_v_suaa`



We have found that colony diameters are significantly greater when sugar
and amino acids are added but that adding sugar alone does not
significantly increase colony diameter.

### Check assumptions

Check the assumptions: All general linear models assume the "residuals"
are normally distributed and have "homogeneity" of variance.

Our first check of these assumptions is to use common sense: diameter is
a continuous and we would expect it to be normally distributed thus we
would expect the residuals to be normally distributed thus we would
expect the residuals to be normally distributed

We then proceed by plotting residuals. The `plot()` function can be used
to plot the residuals against the fitted values (See @fig-anova1-plot1).
This is a good way to check for homogeneity of variance.

:::{#fig-anova1-plot1}

```{r}

plot(mod, which = 1)
```

**Residuals against fitted values.** A plot of the residuals against the 
fitted values allows us to check the homogeneity of variance assumption.
Here, the points are distributed similarly in each group. Any difference 
seems small but perhaps the residuals are more variable for the highest mean.

:::

Perhaps the variance is higher for the highest mean?

We can also use a histogram to check for normality (See
@fig-anova1-plot2).

:::{#fig-anova1-plot2}

```{r}

ggplot(mapping = aes(x = mod$residuals)) + 
  geom_histogram(bins = 8)
```

**Distribution of the residuals**. A histogram of the residuals allows us to 
check the normality assumption. Here, the distribution fairly is symmetrical
and seems consistent with a normal distribution. This is a good sign for 
the assumption of normally distributed residuals.
:::

Finally, we can use the Shapiro-Wilk test to test for normality.

```{r}
shapiro.test(mod$residuals)
```

The p-value is greater than 0.05 so this test of the normality
assumption is not significant.

Taken together, these results suggest that the assumptions of normality
and homogeneity of variance are probably not violated.

### Report

There was a significant effect of media on the diameter of bacterial
colonies (*F* = `r F_value`; *d.f.* = `r df_effect`, `r df_residuals`; 
*p* = `r p_value`). Post-hoc testing
with Tukey's Honestly Significant Difference test [@tukey1949] revealed
the colony diameters were significantly larger when grown with both
sugar and amino acids ($\bar{x} \pm s.e$: `r suaa_m` $\pm$ `r suaa_se` mm) 
than with neither (`r cont_m` $\pm$ `r cont_se` mm; *p* = `r p_cont_v_suaa`) or 
just sugar (`r suga_m` $\pm$ `r suga_se` mm; *p* = `r p_suga_v_suaa`). 
See @fig-culture.

::: {#fig-culture}
```{r}
#| code-fold: true

ggplot() +
  geom_point(data = culture, aes(x = medium, y = diameter),
             position = position_jitter(width = 0.1, height = 0)) +
  geom_errorbar(data = culture_summary, 
                aes(x = medium, ymin = mean - se, ymax = mean + se),
                width = 0.3) +
  geom_errorbar(data = culture_summary, 
                aes(x = medium, ymin = mean, ymax = mean),
                width = 0.2) +
  scale_y_continuous(name = "Diameter (mm)", 
                     limits = c(0, 16.5), 
                     expand = c(0, 0)) +
  scale_x_discrete(name = "Medium", 
                   labels = c("Control", 
                              "Sugar added", 
                              "Sugar and amino acids added")) +
  annotate("segment", x = 2, xend = 3, 
           y = 14, yend = 14,
           colour = "black") +
  annotate("text", x = 2.5,  y = 14.5, 
           label = glue::glue("italic(p) == { p_suga_v_suaa }"),
                              parse = TRUE) +
  annotate("segment", x = 1, xend = 3, 
           y = 15.5, yend = 15.5,
           colour = "black") +
  annotate("text", x = 2,  y = 16, 
           label = glue::glue("italic(p) == { p_cont_v_suaa }"),
                              parse = TRUE) +
  theme_classic()
```

**Medium affects bacterial colony diameter**. Ten replicate colonies
were grown on three types of media: control, with sugar added and with
both sugar and amino acids added. Error bars are means $\pm$ 1 standard
error. There was a significant effect of media on the diameter of
bacterial colonies (*F* = `r F_value`; *d.f.* = `r df_effect`, 
`r df_residuals`; *p* = `r p_value`)). Post-hoc
testing with Tukey's Honestly Significant Difference test [@tukey1949]
revealed the colony diameters were significantly larger when grown with
both sugar and amino acids than with neither or just sugar. Data
analysis was conducted in R [@R-core] with tidyverse packages
[@tidyverse].
:::

# Kruskal-Wallis

Our examination of the assumptions revealed a possible violation of the
assumption of homogeneity of variance. We might reasonably apply a
non-parametric test to this data instead.

The Kruskal-Wallis [@kruskal1952] is non-parametric equivalent of a
one-way ANOVA. The general question you have about your data - do these
groups differ (or does the medium effect diameter) - is the same, but
one of more of the following is true:

-   the response variable is not continuous
-   the residuals are not normally distributed
-   the sample size is too small to tell if they are normally
    distributed.
-   the variance is not homogeneous

Summarising the data using the median and interquartile range is more
aligned to the type of analysis than using means and standard
deviations:

```{r}
culture_summary <- culture |> 
  group_by(medium) |> 
  summarise(median = median(diameter),
            interquartile  = IQR(diameter),
            n = length(diameter))
```

```{r}
#| echo: false
# extract values for ease of reporting
control_m <- culture_summary |> 
  filter(medium == "control") |> 
  pull(median) |> round(1)

control_n <- culture_summary |> 
  filter(medium == "control") |> 
  pull(n) 

suga_m <- culture_summary |> 
  filter(medium == "sugar added") |> 
  pull(median) |> round(1)


suga_n <- culture_summary |> 
  filter(medium == "sugar added") |> 
  pull(n)

suaa_m <- culture_summary |> 
  filter(medium == "sugar and amino acids added") |> 
  pull(median) |> round(1)


suaa_n <- culture_summary |> 
  filter(medium == "sugar and amino acids added") |> 
  pull(n)

```


View the results:

```{r}
culture_summary
```

### Apply `kruskal.test()`

We pass the dataframe and variables to `kruskal.test()` in the same way
as we did for `lm()`. We give the data argument and a "formula" which
says `diameter ~ medium` meaning "explain diameter by medium".

```{r}
kruskal.test(data = culture, diameter ~ medium)

```

```{r}
#| echo: false
#| warning: false
# extract values from mod using broom for ease of reporting
krus_tidy <- broom::tidy(kruskal.test(data = culture, diameter ~ medium))

chisq_value <- krus_tidy |> pull(statistic) |> round(2)
p_value <- krus_tidy |> pull(p.value) |> round(4)
df <- krus_tidy |> pull(parameter) 

```


The result of the test is given on this line:
`Kruskal-Wallis chi-squared = 8.1005, df = 2, p-value = 0.01742`.
`Chi-squared` is the test statistic. The *p*-value is less than 0.05
meaning there is a significant effect of medium on diameter.

Notice that the *p*-value is a little larger than for the ANOVA. This is
because non-parametric tests are generally more conservative (less
powerful) than their parametric equivalents.

A significant Kruskal-Wallis tells us *at least two* of the groups differ
but where do the differences lie? The Dunn test [@dunn1964] is a
post-hoc multiple comparison test for a significant Kruskal-Wallis. It
is available in the package **`FSA`** [@FSA]

Load the package using:

```{r}
library(FSA)
```

Then run the post-hoc test with:

```{r}
dunnTest(data = culture, diameter ~ medium)
```

The `P.adj` column gives *p*-value for the comparison listed in the
first column. `Z` is the test statistic. The *p*-values are a little
larger for the `control - sugar and amino acids added` comparison and
the `sugar added - sugar and amino acids added` comparison but they are
still less than 0.05. This means our conclusions are the same as for the
ANOVA.

```{r}
#| echo: false
#| warning: false
# extract values from dunn for ease of reporting
dunn_tidy <- dunnTest(data = culture, diameter ~ medium)$res

p_cont_v_suga <- dunn_tidy |> 
  filter(Comparison == "control - sugar added") |> 
  pull(P.adj) |> round(4)
p_cont_v_suaa <- dunn_tidy |> 
  filter(Comparison == "control - sugar and amino acids added") |> 
  pull(P.adj) |> round(4)
p_suga_v_suaa <- dunn_tidy |> 
  filter(Comparison == "sugar added - sugar and amino acids added") |> 
  pull(P.adj) |> round(4)

```




### Report

There is a significant effect of media on the diameter of bacterial
colonies (Kruskal-Wallis: *chi-squared* = `r chisq_value`; 
*df* = `r df`; *p*-value = `r p_value`) with colonies growing 
significantly better when both sugar and amino acids are added to the medium. 
Post-hoc testing with the Dunn test [@dunn1964] revealed the colony diameters 
were significantly larger when grown with both sugar and amino acids 
(median = `r suaa_m` mm) than with neither (median = `r cont_m` mm; 
*p* = `r p_cont_v_suaa`) or just sugar (median = `r suga_m` mm;
*p* = `r p_suga_v_suaa`). See @fig-culture-kw.

::: {#fig-culture-kw}
```{r}
#| code-fold: true

ggplot(data = culture, aes(x = medium, y = diameter)) +
 geom_boxplot() +
  scale_y_continuous(name = "Diameter (mm)", 
                     limits = c(0, 16.5), 
                     expand = c(0, 0)) +
  scale_x_discrete(name = "Medium", 
                   labels = c("Control", 
                              "Sugar added", 
                              "Sugar and amino acids added")) +
  annotate("segment", x = 2, xend = 3, 
           y = 14, yend = 14,
           colour = "black") +
  annotate("text", x = 2.5,  y = 14.5, 
           label = glue::glue("italic(p) == { p_suga_v_suaa }"),
                              parse = TRUE) +
    annotate("segment", x = 1, xend = 3, 
           y = 15.5, yend = 15.5,
           colour = "black") +
  annotate("text", x = 2,  y = 16, 
           label = glue::glue("italic(p) == { p_cont_v_suaa }"),
                              parse = TRUE) +
  theme_classic()
```

**Medium affects bacterial colony diameter**. Ten replicate colonies
were grown on three types of media: control, with sugar added and with
both sugar and amino acids added. The heavy lines indicate median
diameter, boxes indicate the interquartile range and whiskers the range.
There was a significant effect of media on the diameter of bacterial
colonies (Kruskal-Wallis: *chi-squared* = `r chisq_value`; 
*df* = `r df`; *p*-value = `r p_value`). Post-hoc testing with the Dunn test 
[@dunn1964] revealed the colony diameters were significantly larger when
grown with both sugar
and amino acids than with neither or just sugar. Data analysis was
conducted in R [@R-core] with tidyverse packages [@tidyverse].
:::

# Summary

1.  A linear model with one explanatory variable with two or more groups
    is also known as a **one-way ANOVA**.

2.  We estimate the **coefficients** (also called the **parameters**) of
    the model. For a one-way ANOVA with three groups these are the mean
    of the first group, $\beta_0$, the difference between the means of
    the first and second groups, $\beta_1$, and the difference between
    the means of the first and third groups, $\beta_2$. We test whether
    the parameters differ significantly from zero

3.  We can use `lm()` to one-way ANOVA in R.

4.  In the output of `lm()` the coefficients are listed in a table in
    the Estimates column. The *p*-value for each coefficient is in the
    test of whether it differs from zero. At the bottom of the output
    there is an $F$ test of the model *overall*. Now we have more than
    two parameters, this is different from the test on any one
    parameter. The R-squared value is the proportion of the variance in
    the response variable that is explained by the model. It tells us is
    the explanatory variable is useful in predicting the response
    variable overall.

5.  When the $F$ test is significant there is a significant effect of
    the explanatory variable on the response variable. To find out which
    means differ, we need a **post-hoc** test. Here we use Tukey’s HSD
    applied with the `emmeans()` and `pairs()` functions from the
    **`emmeans`** package. Post-hoc tests make adjustments to the
    *p*-values to account for the fact that we are doing multiple tests.

6.  The assumptions of the general linear model are that the residuals
    are normally distributed and have homogeneity of variance. A
    residual is the difference between the predicted value and the
    observed value.

7.  We examine a histogram of the residuals and use the Shapiro-Wilk
    normality test to check the normality assumption. We check the
    variance of the residuals is the same for all fitted values with a
    residuals vs fitted plot.

8.  If the assumptions are not met, we can use the Kruskal-Wallis test
    applied with `kruskal.test()` in R and follow it with The Dunn test
    applied with `dunnTest()` in the package **`FSA`**.

9.  When reporting the results of a test we give the significance,
    direction and size of the effect. Our figures and the values we give
    should reflect the type of test we have used. We use means and
    standard errors for parametric tests and medians and interquartile
    ranges for non-parametric tests. We also give the test statistic,
    the degrees of freedom (parametric) or sample size (non-parametric)
    and the p-value. We annotate our figures with the p-value, making
    clear which comparison it applies to.
