# One-way ANOVA and Kruskal-Wallis

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("drafting")
```

## 

More than two samples: One-way ANOVA and Kruskal-Wallis

last chapter

-   Independent and non-independent samples
-   Two-sample-tests – paired and unpaired, parametric and non-parametric

This chapter: Extend our ability to test for differences between two or more groups: one-way ANOVA and its non-parametric equivalent Kruskal-Wallis

-   Why not do several two-sample tests?
-   ANOVA terminology and concepts
-   ANOVA assumptions
-   Running, interpreting and reporting an ANOVA
-   Post-hoc analysis (after a significant ANOVA)
-   When assumptions are not met: Kruskal-Wallis
-   Running, interpreting and reporting Kruskal-Wallis
-   Post-hoc analysis (after a significant Kruskal-Wallis)


Why ANOVA
Type I error: Rejecting the null hypothesis when it is true. This will happen with a probability of 0.05 Doing lots of comparisons increases the type 1 error rate ANOVA tests for an effect of the explanatory variable without increasing type 1 error rate

Assumptions and alternative

ANOVA assumes the “residuals” are normally distributed and have homogeneity of variance
Kruskal-Wallis is the non-parametric equivalent when assumptions are not met


Which growth medium is best for growing bacterial cultures?

Explanatory variable is type of media: categorical with 3 groups
Control
Control + sugar
Control + sugar + amino acids 

Response variable is colony diameters (mm)

One response, one categorical explanatory variable (“one-way ANOVA” or “one-factor ANOVA”)
These data are in tidy format.

[culture.csv](data-raw/culture.csv)

Plot your data roughly

Summarise the data
```{r}
# culture_summary <- culture %>%
#   group_by(medium) %>%
#   summarise(mean = mean(diameter),
#             std = sd(diameter),
#             n = length(diameter),
#             se = std/sqrt(n))
```

ANOVA assumes the “residuals” are normally distributed and have
homogeneity of variance First use common sense: colony diameter is continuous and we would expect it to be normally distributed thus we would expect the residuals to be normally distributed


Reporting the result: “significance, direction, magnitude”

There is a significant effect of media on the diameter of bacterial colonies (ANOVA: F = 6.11; d.f. = 2, 27; p = 0.006).
Or
There is a significant difference in diameters between colonies grown on different media (ANOVA: F = 6.11; d.f. = 2, 27; P=0.006).

What about direction and magnitude??

Which means differ? Post-hoc test needed 
```{r}
library(lsmeans)
library(multcompView)
```

```{r}
# lsmeans(mod, ~ species) %>%
#   pairs()

```

There is a significant effect of media on the diameter of bacterial colonies (F =
6.11; d.f. = 2, 27; p = 0.006) with colonies growing significantly better when both sugar
and amino acids are
added to the medium
(see Figure 1).



```{r}
# ggplot() +
#   geom_jitter(data = seal, 
#               aes(x = species, y = myoglobin), 
#               width = 0.25, colour = "grey") +
#   geom_errorbar(data = seal_summary,
#                 aes(x = species,
#                     ymin = mean,
#                     ymax = mean),
#                 width = .3) +
#   geom_errorbar(data = seal_summary,
#                 aes(x = species,
#                     ymin = mean - se,
#                     ymax = mean + se),
#                 width = .5) +
#   geom_segment(aes(x = 1, y = 71, xend = 2, yend = 71),
#                size = 1) +
#   geom_segment(aes(x = 1, y = 71, xend = 1, yend = 69),
#                size = 1) +
#   geom_segment(aes(x = 2, y = 71, xend = 2, yend = 69),
#                size = 1) +
#   annotate("text", x = 1.5, y = 73,  label = "**", size = 6) +
#   scale_x_discrete(name = "Species") +
#   scale_y_continuous(name = expression("Myoglobin concentration g "*Kg^{-1}),
#                      expand = c(0, 0),
#                      limits = c(0, 75)) +
#   theme_classic()
```

