# Single linear regression {#sec-single-linear-regression}

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("complete")
```

## Overview

Single linear regression is an appropriate way to analyse data when:

-   You have two continuous variables
-   One of the variables is explanatory and the other is a response.
    That is, one variable, the $x$, "causes" the $y$.
-   The explanatory variable has been chosen, set or manipulated and the
    other variable is the measured response. This is sometimes described
    as the $x$ being “sampled without error”
-   The response variable, $y$, is randomly sampled for each $x$ with a
    normal distribution and those normal distributions have the same
    variance.
-   The relationship between the variables is linear

Applying a single linear regression to data means putting a line of best
fit through it. The intercept and the slope of the true population
relationship is estimated from the sample you have. We test whether
those two parameters differ significantly from zero.

### Reporting

Reporting [the significance of effect, direction of effect, magnitude of
effect](what-statistical-model.html#reporting)
for a single linear regression means making the following clear to the
reader:

1.  the significance of effect - whether the slope is significantly
    different from zero

2.  the direction of effect - whether the slope is positive or negative

3.  the magnitude of effect - the slope itself

Figures should reflect what you have said in the statements. Ideally
they should show both the raw data and the statistical model:

We will explore all of these ideas with an example.

## 🎬 Your turn!

If you want to code along you will need to start a new [RStudio
project](workflow-rstudio.html#rstudio-projects), add a `data-raw`
folder and open a new script. You will also need to load the
**`tidyverse`** package [@tidyverse].

## Single linear regression

Three replicate water baths were set up at each of five temperatures
(10, 11C, 12C, 13C, 14C). Ten Brine Shrimp (*Artemia salina*) were
placed in each and their average respiration rate per water bath was
measured (in arbitrary units). The data are in
[shrimp.txt](data-raw/shrimp.txt). In this scenario our null hypothesis,
$H_0$, is that temperature has no effect on respiration rate. Another 
way of saying this is the slope of the line is zero. This is written 
as: $H_0: \beta_1 = 0$. 

### Import and explore

Import the data:

```{r}
shrimp <- read_table("data-raw/shrimp.txt")
```

```{r}
#| echo: false

knitr::kable(shrimp) |> 
  kableExtra::kable_styling() |> 
  kableExtra::scroll_box(height = "200px")
```

These data are in tidy format [@Wickham2014-nl] - all the respiration
values are in one column with another column indicating the water bath
temperature. There is only one water bath per row. This means they are
well formatted for analysis and plotting.

In the first instance, it is sensible to create a rough plot of our data
(See @fig-shrimp-rough). Plotting data early helps us in multiple ways:

-   it helps identify whether there missing or extreme values
-   it allows us to see if the relationship is roughly linear
-   it tells us whether any relationship positive or negative

Scatter plots (`geom_point()`) are a good choice for exploratory
plotting with data like these.

:::{#fig-shrimp-rough}

```{r}
ggplot(data = shrimp,
       aes(x = temperature, y = respiration)) +
  geom_point()
```

**Relationship between temperature and respiration rate in brine shrimp** 
The default scatter plot is enough for us to see that respiration rate seems 
to increase with temperature approximately linearly. 

:::

The figure suggests that respiration rate increases with temperature and
there are no particularly extreme values. We can also see that any
relationship is roughly linear.

### Do a regression with `lm()`

We can create a single linear regression model like this:

```{r}
mod <- lm(data = shrimp, respiration ~ temperature)
```

And examine the model with:

```{r}
summary(mod)
```

```{r}
#| echo: false

b0 <- mod$coefficients[1] |> round(2)
b1 <- mod$coefficients[2] |> round(2)
```

What do all these results mean?

The `Estimate` in the `Coefficients` table give:

-   the `(Intercept)`, known as $\beta_0$, which is the value of *y*
    (the response) when the value of *x* (the explanatory) is zero.

-   the slope, labelled `temperature` known as $\beta_1$, which is the
    amount of *y* to add for each unit of *x*. Since `temperature` is
    positive we know respiration rate *increases* with temperature

@fig-regression-lm-model shows the model and its parameters.

The *p*-values on each line are tests of whether that coefficient is
different from zero. Thus it is:

`temperature  0.91850    0.09182  10.003 1.79e-07 ***`

that tells us the slope is significantly different from zero and thus
there *is* a significant relationship between temperature and
respiration rate.

The *F* value and *p*-value in the last line are a test of whether the
model as a whole explains a significant amount of variation in the
response variable. For a regression, this is exactly equivalent to the
test of the slope against zero and the two *p*-values will be the same.


:::{#fig-regression-lm-model}

```{r}
#| echo: false

ggplot(data = shrimp, 
                aes(x = temperature, y = respiration)) +
  geom_point(colour = "black") +   
  geom_abline(intercept = b0,
              slope = b1,
              pal3[1], linetype = 2) +
  geom_smooth(method = "lm", 
              se = FALSE,
              colour = pal3[1]) +
  scale_x_continuous(expand = c(0,0),
                     limits = c(0, 15),
                      name = "Temperature (C)") +
  scale_y_continuous(expand = c(0,0), 
                     limits = c(-0, 20),
                     name = "Respiration (units)") +
  geom_segment(aes(x = 2,
                   xend = 0,
                   y = b0 + 6,
                   yend = b0),
               colour = pal3[2]) +
  annotate("text",
           x = 3, 
           y = b0 + 6.5,
           label = glue::glue("Intercept (β0) is { b0 }"), 
           colour = pal3[2],
           size = 4) +
  geom_segment(aes(x = 9,
                   xend = 9,
                   y = b0 + b1 * 8,
                   yend = b0 + b1 * 9),
               colour = pal3[3]) +
   geom_segment(aes(x = 8,
                   xend = 9,
                   y = b0 + b1 * 8,
                   yend = b0 + b1 * 8),
               colour = pal3[3]) +
  geom_segment(aes(x = 9,
                   xend = 12,
                   y = b0 + b1 * 8.5,
                   yend = b0 + 1.3),
               colour = pal3[2]) +
  annotate("text",
           x = 12, 
           y = b0 + 1,
           label = glue::glue("slope (β1) is { b1 }"), 
           colour = pal3[2],
           size = 4) +
  theme_classic()

```

**The a linear model fitted to the relationship between temperature and 
respiration rate.** The model has two coefficients (parameters) estimated from 
the data. The first estimate is the intercept and the second estimate is the 
slope.
:::

### Check assumptions

Check the assumptions: All general linear models assume the "residuals"
are normally distributed and have "homogeneity" of variance.

Our first check of these assumptions is to use common sense: respiration
is a continuous variable and we would expect it to be normally
distributed thus we would expect the residuals to be normally
distributed

We then proceed by plotting residuals. The `plot()` function can be used
to plot the residuals against the fitted values (See
@fig-regression-plot1). This is a good way to check for homogeneity of
variance.

:::{#fig-regression-plot1}

```{r}

plot(mod, which = 1)
```

*Residuals against fitted values.** A plot of the residuals against the 
fitted values is a "diagnostic plot" that allows us to check the homogeneity of
variance assumption. Here we see no obvious pattern as the points are 
roughly evenly distributed around the line. This is a good sign for the 
assumption of homogeneity of variance.
:::



We can also use a histogram to check for normality (See
@fig-regression-plot2).

:::{#fig-regression-plot2}

```{r}
ggplot(mapping = aes(x = mod$residuals)) + 
  geom_histogram(bins = 5)
```

**Distribution of the residuals.** A histogram of the residuals is a 
"diagnostic plot" that allows us to check the normality assumption. Here, the
distribution fairly is symmetrical and seems consistent with a normal 
distribution. This is a good sign for the assumption of normally 
distributed residuals."
:::

Finally, we can use the Shapiro-Wilk test to test for normality.

```{r}
shapiro.test(mod$residuals)
```

The p-value is greater than 0.05 so this test of the normality
assumption is not significant. Note that "not significant" means not
significantly different from a normal distribution. It does not mean
definitely normally distributed.

Taken together, these results suggest that the assumptions of normality
and homogeneity of variance are not violated.

### Report

The temperature explained a significant amount of the variation in
respiration rate (ANOVA: *F* = 67; *d.f*. = 1, 13; *p* \< 0.001). The
regression line is: Respiration rate = `r b0` + `r b1` \* temperature.
See @fig-shrimp.

::: {#fig-shrimp}
```{r}
#| code-fold: true

ggplot(data = shrimp, 
                aes(x = temperature, y = respiration)) +
  geom_point(size = 2) +   
  geom_smooth(method = "lm", 
              se = FALSE,
              colour = "black") +
  scale_x_continuous(expand = c(0,0),
                     limits = c(0, 15.5),
                      name = "Temperature (C)") +
  scale_y_continuous(expand = c(0,0), 
                     limits = c(0, 20),
                     name = "Respiration (units)") +
  theme_classic()
```

**Respiration rate of *Artemia salina* increases with temperature**.
Three replicate water baths were set up at each of five temperatures
(10, 11C, 12C, 13C, 14C). Ten Brine Shrimp (*Artemia salina*) were
placed in each and their average respiration rate per water bath was
measured. There was a significant effect of altering water bath
temperature on the average respiration rate (ANOVA: *F* = 67; *d.f*. =
1, 13; *p* \< 0.001). The regression line is: Respiration rate =
`r b0` + `r b1` \* temperature. Data analysis was conducted in R
[@R-core] with tidyverse packages [@tidyverse].
:::

## Summary

1.  Single linear regression is an appropriate when you have one
    continuous explanatory variable and one continuous response and the
    relationship between the two is linear.

2.  Applying a single linear regression to data means putting a line of
    best fit through it. We estimate the **coefficients** (also called 
    the **parameters**) of the model. These are the intercept, $\beta_0$,
    and the slope, $\beta_1$. We test whether the parameters differ
    significantly from zero

3.  We can use `lm()` to a linear regression.

4.  In the output of `lm()` the coefficients are listed in a table in the
    Estimates column. The *p*-value for each coefficient is in the test 
    of whether it differs from zero. At the bottom of the output there 
    is a test of the model *overall*. In a single linear regression this
    is exactly the same as the test of the $\beta_1$ and the p-values are 
    identical. The R-squared value is the proportion of the variance
    in the response variable that is explained by the model.

5.  The assumptions of the general linear model are that the residuals
    are normally distributed and have homogeneity of variance. A
    residual is the difference between the predicted value and the
    observed value.

6.  We examine a histogram of the residuals and use the Shapiro-Wilk
    normality test to check the normality assumption. We check the
    variance of the residuals is the same for all fitted values with a
    residuals vs fitted plot.

7.  If the assumptions are not met, we might need to transform the data
    or use a different type of model.

8.  When reporting the results of a regression we give the significance,
    direction and size of the effect. Often we give the equation of the
    best fitting line. A Figure should show the data and the line of
    best fit.
