# Single linear regression
```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("polishing")
```

## Overview

Single linear regression is an appropriate way to analyse data when:


-   you have two continuous variables
-   there is a cause and effect relationship between the variables
-   the explanatory variable has been chosen, set or manipulated and the other variable is the measured response
-   the relationship between the variables is linear

include principle: var explained by model is greater then residual var. are the parameters different from zero. 

y

### Assumptions

Linear

Prediction: One variable causes the other

Axes cannot be switched

X is “sampled without error”; y randomly sampled for each x

### Reporting

Reporting the result: “significance of effect, direction of effect, magnitude of effect”


In reporting the result of two-sample test we give:

1. the significance of effect - whether there is there an effect 

2. the direction of effect - positive or negative slope

3. the magnitude of effect - slope


Figures should reflect what you have said in the statements. Ideally they should show both the raw data and the statistical model:

We will explore all of these ideas with an example.

## 🎬 Your turn!

If you want to code along you will need to start a new [RStudio project](workflow_rstudio.html#rstudio-projects) then a new script.

## Single linear regression


Three replicates water baths were set up at each of five temperatures (10, 11C, 12C, 13C, 14C) Ten Brine Shrimp (*Artemia salina*) were placed in each and their average respiration rate per water bath was measured (in arbitrary units). The data are in [shrimp.txt](data-raw/shrimp.txt)

### Import and explore

Import the data:

```{r}
shrimp <- read_table("data-raw/shrimp.txt")
```

```{r}
#| echo: false

knitr::kable(shrimp) %>%
  kableExtra::kable_styling() |> 
  kableExtra::scroll_box(height = "200px")
```

These data are in tidy format [@Wickham2014-nl] - all the respiration values are in one column with another column indicating the water bath temperature. There is only one water bath per row. This means they are well formatted for analysis and plotting.

In the first instance, it is sensible to create a rough plot of our data (See @fig-shrimp-rough). Plotting data early helps us in multiple ways:
-   it helps identify whether there missing or extreme values
-   it allows us to see if the relationship is roughly linear
-   it tells us whether any relationship positive or negative 

Scatter plots (`geom_point()`) are a good choice for exploratory plotting with data like these.

```{r}
#| label: fig-shrimp-rough
#| fig-cap: "A default scatter plot of the relationship between temperature and respiration rate in brine shrimp is enough for us to see that respiration rate seems to increase with temperature approximately linearly"

ggplot(data = shrimp,
       aes(x = temperature, y = respiration)) +
  geom_point()
```

The figure suggests that respiration rate increases with temperature and there are no particularly extreme values. We can also see that any relationship is roughly linear. 

### Apply `lm()`

We can create a single linear regression model like this:

```{r}
mod <- lm(data = shrimp, respiration ~ temperature)
```

And examine the model with:

```{r}
summary(mod)
```
```{r}
#| echo: false

b0 <- mod$coefficients[1] |> round(2)
b1 <- mod$coefficients[2] |> round(2)
```


The Estimates in the Coefficients table give:

-   the `(Intercept)` known as $\beta_0$,  which is the value of the *y* (the response) when the value of *x* (the explanatory) is zero. 

-   the slope labelled `temperature` known as $\beta_1$, which is the amount of *y* you add for each unit of *x*. `temperature` is positive so respiration rate increases with temperature

The *p*-values on each line are tests of whether that coefficient is different from zero. Thus it is:

`temperature  0.91850    0.09182  10.003 1.79e-07 ***`

that tells us the relationship is significant.

The *F* value and *p*-value in the last line are a test of whether the model as a whole explains a significant amount of variation in the response variable. For a regression, this is exactly equivalent to the test of the slope against zero and the two *p*-values will be the same.

```{r}
#| echo: false
#| label: fig-regression-lm-model
#| fig-cap: "In a linear model, the first estimate is the intercept and the second estimate is the 'slope'."
ggplot(data = shrimp, 
                aes(x = temperature, y = respiration)) +
  geom_point(colour = "black") +   
  geom_abline(intercept = b0,
              slope = b1,
              pal3[1], linetype = 2) +
  geom_smooth(method = "lm", 
              se = FALSE,
              colour = pal3[1]) +
  scale_x_continuous(expand = c(0,0),
                     limits = c(0, 15),
                      name = "Temperature (C)") +
  scale_y_continuous(expand = c(0,0), 
                     limits = c(-0, 20),
                     name = "Respiration (units)") +
  geom_segment(aes(x = 2,
                   xend = 0,
                   y = b0 + 6,
                   yend = b0),
               colour = pal3[2]) +
  annotate("text",
           x = 3, 
           y = b0 + 6.5,
           label = glue::glue("Intercept (β0) is { b0 }"), 
           colour = pal3[2],
           size = 4) +
  geom_segment(aes(x = 9,
                   xend = 9,
                   y = b0 + b1 * 8,
                   yend = b0 + b1 * 9),
               colour = pal3[3]) +
   geom_segment(aes(x = 8,
                   xend = 9,
                   y = b0 + b1 * 8,
                   yend = b0 + b1 * 8),
               colour = pal3[3]) +
  geom_segment(aes(x = 9,
                   xend = 12,
                   y = b0 + b1 * 8.5,
                   yend = b0 + 1.3),
               colour = pal3[2]) +
  annotate("text",
           x = 12, 
           y = b0 + 1,
           label = glue::glue("slope (β1) is { b1 }"), 
           colour = pal3[2],
           size = 4) +
  theme_classic()

```

### Check assumptions

Check the assumptions: All t-tests assume the "residuals" are normally distributed and have homogeneity of variance.

First use common sense: respiration is a continuous and we would expect it to be normally distributed thus we would expect the residuals to be normally distributed

Second by plotting residuals:

```{r}
plot(mod, which = 1)
```

```{r}
ggplot(mapping = aes(x = mod$residuals)) + 
  geom_histogram(bins = 5)
```

```{r}
shapiro.test(mod$residuals)
```

Checking the assumptions: normally distributed residuals

-   Variance is about the same for all values of x
-   Looks roughly normal, symmetrical
-   test with shapiro. is NS. note that not significant means not significantly different from a normal distribution. It does not mean definitely normally distributed

### Report

The temperature explained a significant amount of the variation in respiration rate (ANOVA: *F* = 67; *d.f*. = 1, 13; *p* < 0.001). The regression line is: Respiration rate = `r b0` + `r b1` * temperature. See @fig-shrimp.

```{r}
#| label: fig-shrimp
#| fig-cap: "The effect of altering water bath temperature on the average respiration rate of brine shrimps."

ggplot(data = shrimp, 
                aes(x = temperature, y = respiration)) +
  geom_point(size = 2) +   
  geom_smooth(method = "lm", 
              se = FALSE,
              colour = "black") +
  scale_x_continuous(expand = c(0,0),
                     limits = c(0, 15.5),
                      name = "Temperature (C)") +
  scale_y_continuous(expand = c(0,0), 
                     limits = c(0, 20),
                     name = "Respiration (units)") +
  theme_classic()
```
