# Two-way ANOVA

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("complete")
```

## Overview

In the last chapter, we learnt how to use and interpret the general
linear model when the *x* variable was categorical with two or more
groups. This procedure is known as one-way ANOVA. We will now
incorporate a *second* categorical variable. This is often known as the
two-way or two-factor ANOVA (**an**alysis **o**f **var**iance). It might
also be described as a "factorial design".

In the last chapter we conducted a one-way ANOVA to test whether there
was an of media on the diameter of bacterial colonies. Suppose we had
run this experiment on more than one bacterial species. We would then
have two explanatory variables: media and species. Perhaps we should do
two one-way ANOVAs, one for each explanatory variable? This would tell
us whether each explanatory variable had an effect on the response
variable. However, it would not tell us whether there was an
**interaction** between the two variables. An interaction is when the
effect of one variable depends on the level of another and in some ways
it is the most interesting result we could reveal. For example, we might
find that the optimal media for one species was not the optimal media
for the other species. A two-way ANOVA allows us to test for the effects
of each explanatory variable *and* whether they interact. A two-way
ANOVA has three null hypotheses:

1.  There is no effect of media on diameter, or there is no difference
    between the mean diameter of colonies grown on different media.
2.  There is no effect of species on diameter, or there is no difference
    between bacteria in colony diameter.
3.  The effect of media is the same for all species, or the effect of
    media does not depend on species.

As a General linear model, two-way ANOVA is a parametric test with
assumptions based on the normal distribution which need to be met for
the *p*-values generated to be accurate. We use `lm()` to fit the model,
`anova()` to test the significance of the explanatory variables and
`emmeans()` and pairs() for post-hoc testing.

### Model assumptions

We examine the model assumptions in the same way as we did for single
linear regression, two-sample tests and one-way ANOVA: we apply the
general linear model with `lm()` and *then* check the assumptions using
diagnostic plots and the Shapiro-Wilk normality test. We also use common
sense:

-   the response should be continuous (or nearly continuous, see [Ideas
    about data: Theory and
    practice](ideas_about_data.html#theory-and-practice)) and we
    consider whether we would expect the response to be normally
    distributed.
-   we expect decimal places and few repeated values.

### Reporting

In reporting the result of two-way ANOVA, we include:

1.  the significance of each of the explanatory variables and the
    interaction between them.

    -   The *F*-statistic and *p*-value for each explanatory variable
        and the interaction term.

2.  the direction of effect - which of the means is greater

    -   Post-hoc test

3.  the magnitude of effect - how big is the difference between the
    means

    -   the means and standard errors for each group

Figures should reflect what you have said in the statements. Ideally
they should show both the raw data and the statistical model: means and
standard errors

### When the assumptions are not met

#### When residuals are not normally distributed

There are a few options for dealing with non-normally distributed
residuals when you wanted to do a two-way ANOVA. Transforming your data
might help - but it depends on how the data are not normal - if there
are lots of values the same for example, no transformation helps! If the
data have positive skew (tail in the positive direction) then logging
can help.

#### When residuals are not normally distributed and/or variance is unequal

There are some rank-based tests, such as the Friedman test and the 
Scheirer-Ray-Hare test, that apply to specific situations. However, 
there is no non-parametric alternative that directly corresponds to 
two-way ANOVA in the same way that the Kruskal-Wallis test does for 
one-way ANOVA.

In these cases you have other good options.

1.  **Use two-way ANOVA carefully**. Using the two-way ANOVA test 
    and interpreting the results with caution is one valid approach. 
    ANOVA is actually quite robust to violating the normality assumption
    (it performs quite well anyway) depending on how much the residuals 
    deviate.
    
2.  **Using non-parametric one-way tests**. Divide the dataset according
    to the groups in one factor, then carry out a Wilcoxon rank-sum test
    (also known as the Mann-Whitney test) or a Kruskal-Wallis test on
    each subset for the other factor. For example, if the two factors
    are Fertilizer Type (Organic, Chemical, Control) and Soil Moisture
    (Low, Medium, High), split the dataset by Fertilizer Type. Within
    each subset, perform a Wilcoxon rank-sum test if the second factor
    has two levels or a Kruskal-Wallis test if it has more than two
    levels to determine whether the second factor affects the response
    variable. While there is no direct test for an interaction, you can
    infer one. If the results are consistent across all subsets, meaning
    the second factor has the same effect regardless of the first
    factor, there is no evidence of an interaction. However, if the
    effect of the second factor varies between subsets, this suggests an
    interaction, as the influence of one factor depends on the level of
    the other.

<!-- 2.  **Using a Permutation Test**. A permutation test checks if an  -->
<!--     observed difference is real or just due to chance by shuffling  -->
<!--     the data many times and seeing how often a difference as big as  -->
<!--     the original one appears. Instead of using formulas based on  -->
<!--     assumptions, it creates a null distribution from the shuffled  -->
<!--     data and calculates a p-value based on how rare the observed  -->
<!--     result is. This makes it useful when normal statistical methods  -->
<!--     don’t apply. -->



We will explore all of these ideas with some examples.

## 🎬 Your turn!

If you want to code along you will need to start a new [RStudio
project](workflow_rstudio.html#rstudio-projects), add a `data-raw`
folder and open a new script. You will also need to load the
**`tidyverse`** package [@tidyverse].

## Two-way ANOVA

Researchers have collected live specimens of two species of periwinkle
(See @fig-periwinkles) from sites in northern England in the Spring and
Summer. They took a measure of the gut parasite load by examining a
slide of gut contents. The data are in
[periwinkle.txt](data-raw/periwinkle.txt). The data were collected to
determine whether there was an effect of season or species on parasite
load and whether these effects were independent.

![Periwinkles are marine gastropod molluscs (slugs and snails). A)
*Littorina brevicula* (PD files - Public Domain,
https://commons.wikimedia.org/w/index.php?curid=30577419) B) *Littorina
littorea*. (photographed by Guttorm Flatabø (user:dittaeva). -
Photograph taken with an Olympus Camedia C-70 Zoom digital camera.
Metainformation edited with Irfanview, possibly cropped with jpegcrop.,
CC BY-SA 3.0,
https://commons.wikimedia.org/w/index.php?curid=324769](images/Littorina.jpg){#fig-periwinkles
fig-alt=""}

### Import and explore

Import the data:

```{r}
periwinkle <- read_delim("data-raw/periwinkle.txt", delim = "\t")
```

```{r}
#| echo: false

knitr::kable(periwinkle) |> 
  kableExtra::kable_styling() |> 
  kableExtra::scroll_box(height = "200px")
```

The Response variable is parasite load and it appears to be continuous.
The Explanatory variables are species and season and each has two
levels. It is known “two-way ANOVA” or “two-factor ANOVA” because there
are two explanatory variables.

These data are in tidy format [@Wickham2014-nl] - all the parasite load
values are in one column (`para`) with the other columns indicating the
`species` and the `season`. This means they are well formatted for
analysis and plotting.

In the first instance it is sensible to create a rough plot of our data.
This is to give us an overview and help identify if there are any issues
like missing or extreme values. It also gives us idea what we are
expecting from the analysis which will make it easier for us to identify
if we make some mistake in applying that analysis.

Violin plots (`geom_violin()`, see @fig-periwinkle-rough), box plots
(`geom_boxplot()`) or scatter plots (`geom_point()`) all make good
choices for exploratory plotting and it does not matter which of these
you choose.

```{r}
#| label: fig-periwinkle-rough
#| fig-cap: "The parasite load for two species of Littorina indicated by the fill colour, in the Spring and Summer. Parasite load seems to be higher for both species in the summer and that effect looks bigger in L.brevicula - it has the lowest spring mean but the highest summer mean."

ggplot(data = periwinkle, 
       aes(x = season, y = para, fill = species)) +
  geom_violin()
```

R will order the groups alphabetically by default.

The figure suggests that parasite load is higher for both species in the
summer and that effect looks bigger in *L.brevicula* - it has the lowest
spring mean but the highest summer mean.

Summarising the data for each species-season combination is the next
sensible step. The most useful summary statistics are the means,
standard deviations, sample sizes and standard errors. I recommend the
`group_by()` and `summarise()` approach:

```{r}
peri_summary <- periwinkle |>  
  group_by(season, species) |>  
  summarise(mean = mean(para),
            sd = sd(para),
            n = length(para),
            se = sd / sqrt(n))
```

We have save the results to `peri_summary` so that we can use the means
and standard errors in our plot later.

```{r}
peri_summary
```

The summary confirms both species have a higher mean in the summer and
that the difference between the species is reversed - *L.brevicula*
minus *L.littorea* is `r peri_summary$mean[1]-peri_summary$mean[2]` in
the spring but `r peri_summary$mean[3]-peri_summary$mean[4]` in summer.

### Apply `lm()`

We can create a two-way ANOVA model like this:

```{r}
mod <- lm(data = periwinkle, para ~ species * season)
```

And examine the model with:

```{r}
summary(mod)
```

The Estimates in the Coefficients table give:

-   `(Intercept)` known as $\beta_0$. This the mean of the group with
    the first level of *both* explanatory variables, *i.e.*, the mean of
    *L.brevicula* group in Spring. Just as the intercept is the value of
    *y* (the response) when the value of *x* (the explanatory) is zero
    in a simple linear regression, this is the value of `para` when both
    `season` and `species` are at their first levels. The order of the
    levels is alphabetical by default.

-   `speciesLittorina littorea` known as $\beta_1$. This is what needs
    to be added to the *L.brevicula* Spring mean (the intercept) when
    the `species` goes from its first level to its second. That is, it
    is the difference between the *L.brevicula* and *L.littorea* means
    in Spring. The `speciesLittorina littorea` estimate is positive so
    the the *L.littorea* mean is higher than the *L.brevicula* mean in
    Spring. If we had more species we would have more estimates
    beginning `species...` and all would be comparisons to the
    intercept.

-   `seasonSummer` known as $\beta_2$. This is what needs to be added to
    the *L.brevicula* Spring mean (the intercept) when the `season` goes
    from its first level to its second. That is, it is the difference
    between the Spring and Summer means for *L.brevicula*. The
    `seasonSummer` estimate is positive so the the Summer mean is higher
    than the Spring mean. If we had more seasons we would have more
    estimates beginning `season...` and all would be comparisons to the
    intercept.

-   `speciesLittorina littorea:seasonSummer` known as $\beta_3$. This is
    interaction effect. It is an additional effect. Going from
    *L.brevicula* to *L.littorea* adds $\beta_1$ to the intercept. Going
    from Spring to Summer adds $\beta_2$ to the intercept. Going from
    *L.brevicula* in Spring to *L.littorea* in Summer adds
    $\beta_1 + \beta_2 + \beta_3$ to the intercept. If $\beta_3$ is zero
    then the effect of `species` is the same in both `season`s. If
    $\beta_3$ is not zero then the effect of `species` is different in
    the two `season`s.

The *p*-values on each line are tests of whether that coefficient is
different from zero.

The *F* value and *p*-value in the last line are a test of whether the
model as a whole explains a significant amount of variation in the
response variable. The model of season and species overall explains a
significant amount of the variation in parasite load
(`p-value: 3.043e-06`). To see which of the three effects are
significant we can use the `anova()` function on our model.

Determine which effects are significant:

```{r}
anova(mod)
```

The parasite load is significantly greater in Summer (*F* = 25.9; *d.f.*
= 1, 96; *p* \< 0.0001) but this effect differs between species (*F* =
6.2; *d.f.* = 1,96; *p* = 0.014) with a greater increase in parasite
load in *L.brevicula* than in *L.littorea*.

We need a post-hoc test to see which comparisons are significant and can
again use then **`emmeans`** [@emmeans] package.

Load the package

```{r}
library(emmeans)
```

Carry out the post-hoc test

```{r}
emmeans(mod, ~ species * season) |> pairs()

```

Each row is a comparison between the two means in the 'contrast' column.
The 'estimate' column is the difference between those means and the
'p.value' indicates whether that difference is significant.

A plot can be used to visualise the result of the post hoc which can be
especially useful when there are very many comparisons.

Plot the results of the post-hoc test:

```{r}
emmeans(mod, ~ species * season) |> plot()
```

We have significant differences between:

-   *L.brevicula* in the Spring and Summer `p <.0001`
-   *L.brevicula* in the Spring and *L.littorea* in the Summer
    `p = 0.0004`
-   *L.littorea* in the Spring *L.brevicula* in the Summer `p = 0.0172`

### Check assumptions

Check the assumptions: All general linear models assume the "residuals"
are normally distributed and have "homogeneity" of variance.

Our first check of these assumptions is to use common sense: diameter is
a continuous and we would expect it to be normally distributed thus we
would expect the residuals to be normally distributed thus we would
expect the residuals to be normally distributed

We then proceed by plotting residuals. The `plot()` function can be used
to plot the residuals against the fitted values (See @fig-anova2-plot1).
This is a good way to check for homogeneity of variance.

```{r}
#| label: fig-anova2-plot1
#| fig-cap: "A plot of the residuals against the fitted values shows whether the points are distributed similarly in each group. Any difference seems small but perhaps the residuals are less variable for the lowest mean."

plot(mod, which = 1)
```

We can also use a histogram to check for normality (See
@fig-anova2-plot2).

```{r}
#| label: fig-anova2-plot2
#| fig-cap: "A histogram of residuals is symetrical and seems consistent with a normal distribution. This is a good sign for the assumption of normally distributed residuals."
ggplot(mapping = aes(x = mod$residuals)) + 
  geom_histogram(bins = 10)
```

Finally, we can use the Shapiro-Wilk test to test for normality.

```{r}
shapiro.test(mod$residuals)
```

The p-value is greater than 0.05 so this test of the normality
assumption is not significant.

Taken together, these results suggest that the assumptions of normality
and homogeneity of variance are probably not violated.

### Report

We might report this result as:

The parasite load was significantly greater in Summer (*F* = 25.9;
*d.f.* = 1, 96; *p* \< 0.0001) but this effect differed between species
(*F* = 6.2; *d.f.* = 1,96; *p* = 0.014) with a greater increase in
parasite load in *L.brevicula* (from $\bar{x} \pm s.e$: 57.0 $\pm$ 1.77
units to 73.5 $\pm$ 2.24 units) than in *L.littorea* (from 64.2 $\pm$
2.38 units to 69.9 $\pm$ 2.30 units). See @fig-para.

::: {#fig-para}
```{r}
#| code-fold: true

ggplot() +
  geom_point(data = periwinkle, aes(x = season,
                                    y = para,
                                    shape = species),
             position = position_jitterdodge(dodge.width = 1,
                                             jitter.width = 0.3,
                                             jitter.height = 0),
             size = 3,
             colour = "gray50") +
  geom_errorbar(data = peri_summary, 
                aes(x = season, ymin = mean - se, ymax = mean + se, group = species),
                width = 0.5,
                linewidth = 1,
                position = position_dodge(width = 1)) +
  geom_errorbar(data = peri_summary, 
                aes(x = season, ymin = mean, ymax = mean, group = species),
                width = 0.4,
                linewidth = 1,
                position = position_dodge(width = 1) ) +
  scale_x_discrete(name = "Season") +
  scale_y_continuous(name = "Number of parasites",
                     expand = c(0, 0),
                     limits = c(0, 140)) +
  scale_shape_manual(values = c(19, 1),
                     name = NULL,
                     labels = c(bquote(italic("L.brevicula")),
                                bquote(italic("L.littorea")))) +
  # *L.brevicula* in the Spring and Summer `p <0.0001`
  annotate("segment",
           x = 0.75, xend = 1.75,
           y = 115, yend = 115,
           colour = "black") +
  annotate("text",
           x = 1.25,  y = 119,
           label = "p < 0.0001") +
  # # *L.brevicula* in the Spring and  *L.littorea* in the Summer `p = 0.0004`
  annotate("segment",
           x = 0.75, xend = 2.25,
           y = 125, yend = 125,
           colour = "black") +
  annotate("text", x = 1.5,  y = 129,
           label = "p = 0.0004") +
  # *L.littorea* in the Spring *L.brevicula* in the Summer `p = 0.0172`
  annotate("segment",
           x = 1.25, xend = 1.75,
           y = 105, yend = 105,
           colour = "black") +
  annotate("text", x = 1.5,  y = 109,
           label = "p = 0.0172") +
  theme_classic() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.9, 0.1),
        legend.background = element_rect(colour = "black"))
```

**Parasite load is greater in Summer especially for *L.brevicula*.**
Live specimens of two species of periwinkle, *L.brevicula* and
*L.littorea* were collected from sites in northern England in the Spring
and Summer and gut parasite load was determined. Error bars are means
$\pm$ 1 standard error. Parasite load was significantly greater in
Summer (*F* = 25.9; *d.f.* = 1, 96; *p* \< 0.0001) but this effect
differed between species (*F* = 6.2; *d.f.* = 1,96; *p* = 0.014) with a
greater increase in parasite load in *L.brevicula* than in *L.littorea*.
Data analysis was conducted in R [@R-core] with tidyverse packages
[@tidyverse]. Post-hoc analysis was carried out with Tukey's Honestly
Significant Difference test [@tukey1949].
:::

## Summary

1.  A linear model with two explanatory variable with two or more groups
    each is also known as a **two-way ANOVA**.

2.  We estimate the **coefficients** (also called the **parameters**) of
    the model. For a two-way ANOVA with two groups in each explanatory
    variable these are

    -   the mean of the group with the first level of both explanatory
        variables $\beta_0$,
    -   what needs to be added to $\beta_0$ when one of the explanatory
        variables goes from its first level to its second, $\beta_1$
    -   what needs to be added to $\beta_0$ when the other of the
        explanatory variables goes from its first level to its second,
        $\beta_2$
    -   $\beta_3$, the interaction effect: what additionally needs to be
        added to $\beta_0$ + $\beta_1$ + $\beta_2$ when both of the
        explanatory variables go from their first level to their second

3.  We can use `lm()` to two-way ANOVA in R.

4.  In the output of `lm()` the coefficients are listed in a table in
    the Estimates column. The *p*-value for each coefficient is in the
    test of whether it differs from zero. At the bottom of the output
    there is an $F$ test of the model *overall*. The R-squared value is
    the proportion of the variance in the response variable that is
    explained by the model.

5.  To see which of the three effects are significant we can use the
    `anova()` function on our model.

6.  To find out which means differ, we need a **post-hoc** test. Here we
    use Tukey’s HSD applied with the `emmeans()` and `pairs()` functions
    from the **`emmeans`** package. Post-hoc tests make adjustments to
    the *p*-values to account for the fact that we are doing multiple
    tests.

7.  The assumptions of the general linear model are that the residuals
    are normally distributed and have homogeneity of variance. A
    residual is the difference between the predicted value and the
    observed value.

8.  We examine a histogram of the residuals and use the Shapiro-Wilk
    normality test to check the normality assumption. We check the
    variance of the residuals is the same for all fitted values with a
    residuals vs fitted plot.

9.  When reporting the results of a test we give the significance,
    direction and size of the effect. We use means and standard errors
    for parametric tests like two-way ANOVA. We also give the test
    statistic, the degrees of freedom (parametric) or sample size
    (non-parametric) and the p-value. We annotate our figures with the
    p-value, making clear which comparison it applies to.
