# What is a statistical model? {#sec-what-statistical-model}

```{r}
#| results: "asis"
#| echo: false

source("_common.R")
status("complete")
```

## Overview

This section discusses statistical models which are equations
representing relationships between variables. Statistical models help us
test hypotheses and make predictions. The process involves estimating
model "parameters" from data and assessing "model fit". Linear models
include regression, *t*-tests, and ANOVA, known collectively as the
General Linear Model. The assumptions of the general linear model are
that the "residuals" are normally distribution and variance is
homogeneous. If the assumptions are violated we can use non-parametric
tests.

## What is a statistical model?

A statistical model is a mathematical equation that helps us understand
the relationships between variables. We evaluate how well our data fit a
particular model so we can infer something about how the values arose or
make predictions about future values.

The equation states what has to be done to the explanatory values to get
the response value. For example, a simple model of plant growth might be
that a plant grows by 2 cm per day for from an age of two weeks old.
This model would be written as:

$$
h_{(t)} = h_{(0) }+ 2t
$$ {#eq-plantgrowth-linear}

Where:

-   $t$ is the time in days after the plant is two weeks old
-   $h_{(t)}$ is the height of the plant at time $t$
-   $h_{(0)}$ is the height of the plant at $t=0$, *i.e.*, at two weeks

This model is a linear model because the relationship between the
response variable, height, and the explanatory variable, time, is linear
(See @fig-model-linear). In a linear model, the gradient of the line is
the same no matter what the value of $t$. In this case, it is fixed at 2
cm per day.

One alternative is a simple exponential model. In an exponential model,
the height might increase by 12% each day and the gradient of the line
would increase over time. This model is written as:

$$
h_{(t)} = h_{(0)}1.12^t
$$

Where:

-   $t$ is the time in days after the plant is two weeks old
-   $h_{(t)}$ is the height of the plant at time $t$
-   $h_{(0)}$ is the height of the plant at $t=0$, *i.e.*, at two weeks

This model is not a straight line (See @fig-model-exponential). The
gradient of the line increase as time goes on.

::: {#fig-model layout-ncol="2"}
```{r}
#| echo: false
#| label: fig-model-linear

b0 <- 3
b1 <-  2
df <- data.frame(days = 0:20)
df <- df |> mutate(height = b0 + b1 * days)
ggplot(df, aes(x = days, y = height)) +
  geom_line(colour = pal3[1], linewidth = 1) +
  scale_x_continuous(name = "Time (days after 2 weeks)",
                     expand = c(0, 0),
                     limits = c(0, 20)) +
  scale_y_continuous(name = "Height (cm)",
                     expand = c(0, 0),
                     limits = c(0, 45)) +
  ggtitle(expression(h[(t)]~"="~h[(0)]~+~2*t)) +
  theme_classic()
```

```{r}
#| echo: false
#| label: fig-model-exponential
b0 <- 3
b1 <-  1.12
df <- data.frame(days = 0:20)
df <- df |> mutate(height = b0 * b1^days)
ggplot(df, aes(x = days, y = height)) +
  geom_line(colour = pal3[1], linewidth = 1) +
  scale_x_continuous(name = "Time (days after 2 weeks)",
                     expand = c(0, 0),
                     limits = c(0, 20)) +
  scale_y_continuous(name = "Height (cm)",
                     expand = c(0, 0),
                     limits = c(0, 45)) +
  ggtitle(expression(h[(t)]~"="~h[(0)]*1.12^t)) +
  theme_classic()

```

**Two possible models of plant growth.** [-@fig-model-linear] shows a
linear model of plant growth where the height increases by a fixed
amount each day. [-@fig-model-exponential] shows an exponential model of
plant growth where the height increases by a fixed percentage each day.
The linear model has a constant gradient, while the exponential model
has a gradient that increases over time.
:::

## Statistical Models and Hypothesis Testing

When we conduct statistical analysis, we are using a statistical model
—a mathematical framework that describes the relationship between
explanatory and response variables. Every model makes some assumptions
about this relationship. We estimate the parameters of the model from
the data. For example, in a simple linear model of plant growth, the key
parameters are the intercept and the slope.

Statistical hypothesis testing is directly tied to these models. When we
test a hypothesis, we are often assessing whether a particular model
parameter is significantly different from zero. This is equivalent to
testing whether an explanatory variable has an effect on the response
variable.

In the context of hypothesis testing, the null hypothesis ($H_0$)
typically states that a given parameter is equal to zero, meaning there
is no effect or no relationship. The alternative hypothesis ($H_1$)
suggests that the parameter is different from zero, indicating an
effect.

To evaluate this, we calculate the probability of obtaining our
estimated parameter value, or a more extreme value, *if the true
parameter were actually zero* (*i.e.*, if $H_0$ were true). This
probability is the *p*-value.

### Assumptions and Model Fit

Because hypothesis tests depend on probability calculations, they also
depend on certain assumptions about the data. A key assumption is that
the estimated parameter follows a normal distribution, which allows us
to use standard probability models to determine significance. This
assumption is reasonable in many cases, but if it is violated, the
*p*-values may be misleading.

For this reason, checking model assumptions is crucial. If the
assumptions are not met, we might need to transform the data, choose a
different model, or use **non-parametric tests**, which make fewer
assumptions about data distribution. Non-parametric methods do not
estimate parameters (like an intercept and slope), which makes them
useful when standard parametric assumptions are questionable.

## Using a linear model in practice.

Imagine we are studying a population of bacteria and want understand how
nutrient availability influences its growth. We could grow the bacteria
with different levels of nutrients and measure the diameter of bacterial
colonies on agar plates in a controlled environment so that everything
except the nutrient availability was identical. We could then plot the
diameters against the nutrient levels.

We might expect the relationship between nutrient level and growth to be
linear and add a line of best fit. See @fig-linear-bact.

```{r}
#| echo: false
set.seed(123)
df1 <- data.frame(Nutrient = seq(0.1, 3, 0.1))
df1 <- df1 |> 
  mutate(Diameter = (3 + Nutrient * 2) + rnorm(length(Nutrient)))
mod <- lm(data = df1, Diameter ~ Nutrient)
b0 <- mod$coefficients[1] |> round(2)
b1 <- mod$coefficients[2] |> round(2)

```

::: {#fig-linear-bact layout-ncol="2"}
```{r}
#| echo: false
#| label: fig-linear-bact-without

ggplot(df1, aes(x = Nutrient, y = Diameter)) +
  geom_point() +
  scale_x_continuous(expand = c(0, 0),
                     limits = c(0, 3.2)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 12)) +
  theme_classic()
```

```{r}
#| echo: false
#| label: fig-linear-bact-with
ggplot(df1, aes(x = Nutrient, y = Diameter)) +
  geom_point() +
  scale_x_continuous(expand = c(0, 0),
                     limits = c(0, 3.2)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 12)) +
  geom_smooth(method = "lm", 
              se = FALSE,
              colour = "black") +
  geom_text(label = glue::glue("Diameter = {b0} + {b1} * Nutrient"),
            x = 1, y = 11) +
  theme_classic()

```

**The effect of nutrient level on bacterial colony diameters**.
[-@fig-linear-bact-without] Data plotted without and
[-@fig-linear-bact-with] with a linear model.
:::

The equation of this line is a statistical model that allows us to make
predictions about colony diameter from nutrient levels. A line - or
linear model - has the form:

$$
y = \beta_{0} + \beta_{1}x
$$ {#eq-linear-model}

Where:

-   $y$ is the response variable and $x$ is the explanatory variable.
-   $\beta_{0}$ is the value of $y$ when $x = 0$ usually known as the
    intercept
-   $\beta_{1}$ is the amount added to $y$ for each unit increase in $x$
    usually known as the slope

$\beta_{0}$ and $\beta_{1}$ are called the *coefficients* -- or
*parameters* -- of the model.

In this case $$
Diameter = \beta_{0} + \beta_{1}Nutrient
$$ {#eq-linear-model-bact}

Linear models are amongst the most commonly used statistics. Regression,
*t*-tests and ANOVA are all linear models collectively known as the
*General Linear Model*.

## Model fitting

Fitting a linear model is the process of estimating the parameters
$\beta_{0}$ and $\beta_{1}$ from data. The line gives the values of $y$
which the model predicts. For any value of $x$, the actual measured
value of $y$ will differ from the predicted value and this difference is
called a *residual* or an *error*. The line is a best fit in the sense
that $\beta_{0}$ and $\beta_{1}$ minimises the *sum of the squared
residuals*. The sum of the squared residuals is usually abbreviated
$SSE$.

$$
SSE = \sum(y_{i}-\hat{y})^2
$$ {#eq-sse}

Where:

-   $y_{i}$ represents each of the measured $y$ values from the 1st to
    the *i*th
-   $\hat{y}$ is predicted value

Since $\beta_{0}$ and $\beta_{1}$ are those that minimise the $SSE$,
they are described as *least squares estimates*. You do not need to
worry about this too much but it is a useful piece of statistical jargon
to have heard of because it pops up often. The mean of a sample is also
a least squares estimate because the sum of the squared differences
between each value and the mean is smaller than the sum of the squared
differences between each value and any other value.

The role played by $SSE$ in estimating our parameters means that it is
also used in determining how well our model fits our data. Our model can
be considered useful if the difference between the actual measured value
of $y$ and the predicted value is small but $SSE$ will also depend on
the size of $y$ and the sample size. This means we express $SSE$ as a
*proportion* of the total variation in $y$. The total variation in $y$
is denoted $SST$:

$$
\frac{SSE}{SST}
$$ {#eq-residual-var}

$\frac{SSE}{SST}$ is called the *residual variation*. It is the
proportion of variance remaining after the model fitting. In contrast,
the proportion of the total variance that is explained by the model is
called R-squared, $R^2$. It is:

$$
R^2=1-\frac{SSE}{SST}
$$ {#eq-r-squared}

If there were no explanatory variables, the value we would predict for
the response variable is its mean. In other words, if you did not know
the nutrient level for a randomly chosen bacterial colony the best guess
you could make for its eventual diameter is the mean diameter. Thus, a
good model should fit the response better than the mean - that is, a
good model should fit the response better than a best guess. The output
of `lm()` includes the $R^2$. It represents the proportional improvement
in the predictions from the regression model relative to the mean model.
It ranges from zero (the model is no better than the mean) to 1 (the
predictions are perfect). See @fig-lm-fit

::: {#fig-lm-fit}
![](images/lm_fit.svg)

**A linear model with different fits**. A) the model is a poor fit - the
explanatory variable is no better than the response mean for predicting
the response. B) the model is good fit - the explanatory variable
explains a high proportion of the variance in the response. C) the model
is a perfect fit - the response can be predicted perfectly from the
explanatory variable. Measured response values are in pink, the
predictions are in green and the dashed blue line gives the mean of the
response.
:::

Since the distribution of the responses for a given $x$ is assumed to be
normal and the variances of those distributions are assumed to be
homogeneous, both are also true of the residuals. It is our examination
of the residuals which allows us to evaluate whether the assumptions are
met.

See @fig-lm-annotated for a graphical representation of linear modelling
terms introduced so far.

::: {#fig-lm-annotated}
![](images/generic_lm.svg)

**A linear model annotated with the terms used in modelling**. Measured
response values are in pink, the predictions are in green, and the
differences between these, known as the residuals, are in blue. The
estimated model parameters, $\beta_{0}$ (the intercept) and $\beta_{1}$
(the slope) are indicated.
:::

### General linear model assumptions

The assumptions of the general linear model are that the residuals are
normally distributed and have homogeneity of variance. A residual is the
difference between the predicted and observed value

If we have a continuous response and a categorical explanatory variable
with two groups, we usually apply the general linear model with `lm()`
and then check the assumptions, however, we can sometimes tell when a
non-parametric test would be more appropriate before that:

-   Use common sense - the response should be continuous (or nearly
    continuous, see [Ideas about data: Theory and
    practice](ideas-about-data.html#theory-and-practice)). Consider
    whether you would expect the response to be continuous
-   We would expect decimal places and few repeated values.

To examine the assumptions after fitting the linear model, we plot the
residuals and test them against the normal distribution

## Choice of model

Choosing a statistical model is usually iterative. We first consider
whether a model is appropriate to answer the research and whether the
type of relationship between variables could be expected to be captured
by that type of model. We then fit the model to the data and check the
assumptions. If the assumptions are met, we can proceed with the model.
If the assumptions are not met, we may need to consider a different
model or a non-parametric test.

The consequence of using the wrong model is that your research question
is not answered and *p*-values are not accurate. This can lead to false
conclusions.

## General linear models in R

We use the `lm()` function in R to analyse data with the general linear
model. When you have one explanatory variable the command is:

<code> lm(data = *dataframe*, *response* \~ *explanatory*) </code>

The `response ~ explanatory` part is known as the model **formula**.
These must be the names of two columns in the dataframe.

When you have two explanatory variable we add the second explanatory
variable to the formula using a `+` or a `*`. The command is:

<code> lm(data = *dataframe*, *response* \~ *explanatory1* +
*explanatory2*) </code>

or

<code> lm(data = *dataframe*, *response* \~ *explanatory1* \*
*explanatory2*) </code>

A model with `explanatory1 + explanatory2` considers the effects of the
two variables independently. A model with `explanatory1 * explanatory2`
considers the effects of the two variables *and* any interaction between
them. You will learn more about independent effects and interactions in
[Two-way ANOVA](two_way_anova.html)

We usually assign the output of an `lm()` command to an object and view
it with `summary()`. The typical workflow would be:

<code> mod \<- lm(data = *dataframe*, *response* \~ *explanatory*)\
summary(mod) </code>

There are two sorts of statistical tests in the output of
`summary(mod)`:

1.  tests of whether each coefficient is significantly different from
    zero and,
2.  an *F*-test of the model fit overall

The *F*-test in the last line of the output indicates whether the
relationship modelled between the response and the set of explanatory
variables is statistically significant. *i.e.*, whether it explains a
significant amount of variation.

## Checking assumptions

The assumptions relate to the type of relationship chosen and the
hypothesis testing about the parameters. For a general linear model we
assume the relationship between diameter and nutrients is linear and we
examine this by plotting our data before running any tests.

The assumptions of the hypothesis testing in a general linear model are
that residuals are normally distributed and have homogeneity of
variance. A residual is the difference between the predicted and
observed value. We usually check these assumptions after fitting the
linear model by using the `plot()` function. This produces diagnostic
plots to explore the distribution of the residuals. These cannot prove
the assumptions are met but allow us to quickly determine if the
assumptions are plausible, and if not, how the assumptions are violated
and what data points contribute to the violation.

```{r}
#| echo: false

set.seed(1238)
b0 <- 4
b1 <- 2.5
n <- 100
x <- rnorm(n, mean = 10, sd = 2)
y <-  (b0 + b1 * x) + rnorm(n)
df <- data.frame(x = x, y = y)
mod <- lm(data = df, x ~ y)

```

The two diagnostic plots which are most useful are the "Q-Q" plot (plot
2) and the "Residuals vs Fitted" plot (plot 1). These are given as
values to the `which` argument of `plot()`.

### The Q-Q plot

The Q-Q plot is a scatterplot of the residuals (standardised to a mean
of zero and a standard deviation of 1) against what is expected if the
residuals are normally distributed.

```{r}
plot(mod, which = 2)
```

The points should fall roughly on the line if the residuals are normally
distributed. In the example above, the residuals appear normally
distributed.

The following are two examples in which the residuals are not normally
distributed.

```{r}
#| echo: false

set.seed(100)
y <-  (b0 + b1 * x) + rexp(n,rate = 5)
df <- data.frame(x = x, y = y)
mod2 <- lm(data = df, x ~ y)
plot(mod2, which = 2)
```

If you see patterns like these you should find an alternative to a
general linear model such as a non-parametric test or a generalised
linear model. Sometimes, applying a transformation to the response
variable will result in better meeting the assumptions.

### The Residuals vs Fitted plot

```{r}
#| echo: false

set.seed(1234)
y <-  (b0 + b1 * x) + rpois(n,lambda = 1)
df <- data.frame(x = x, y = y)
mod3 <- lm(data = df, x ~ y)
plot(mod3, which = 2)
```

The Residuals vs Fitted plot shows if residuals have homogeneous
variance or non-linear patterns. Non-linear relationships between
explanatory variables and the response will usually show in this plot if
the model does not capture the non-linear relationship. For the
assumptions to be met, the residuals should be equally spread around a
horizontal line as they are here:

```{r}
plot(mod, which = 1)
```

The following are two examples in which the residuals do not have
homogeneous variance and display non-linear patterns.

```{r}
#| echo: false

set.seed(100)
y <-  (b0 + b1 * x) + rexp(n,rate = 0.1)
df <- data.frame(x = x, y = y)
mod2 <- lm(data = df, x ~ y)
plot(mod2, which = 1)
plot(mod3, which = 1)
```

## Reporting

When reporting the results of statistical tests we need to make sure we
tell the reader everything they need to know and give the evidence it to
support it. What they need to know is given with statements describing
what difference or effect is significant. The evidence is comprises the
test statistic and *p*-value from the test. You can think the
statistical test values as being the evidence for the statements in your
results sections, just as citations are the evidence for the statements
in your introduction.

In reporting the result of a test we give:

1.  the significance of effect

2.  the direction of effect

3.  the magnitude of effect

Figures should demonstrate the statement. Ideally they will include all
the data and the 'model', *i.e.*, the means and error bars or the fitted
line. Figure legends should be concise but contain all the information
needed to understand the figure. I like this blog on [How to craft a
figure legend for scientific
papers](https://bioturing.medium.com/how-to-craft-a-figure-legend-for-scientific-papers-b26a77fa890)

## Summary

1.  A statistical model is an equation that describes the relationship
    between a response variable and one or more explanatory variables.

2.  A statistical model allows you to make predictions about the
    response variable based on the values of the explanatory variables.

3.  Many statistical tests are types of "General Linear Model" including
    linear regression, *t*-tests and ANOVA.

4.  Statistical testing means estimating the model "parameters" and
    testing whether they are significantly different from zero. The
    parameters, also known as coefficients, are the intercept and
    slope (s) in a General Linear Model. A *p*-value less than 0.05 for
    the slope means there is a significant relationship between the
    response and the explanatory variable.

5.  We also consider the fit of the model to the data using the
    *R*-squared value and the *F*-test. An *R*-squared value close to 1
    indicates a good fit and *p*-value less than 0.05 for the *F*-test
    indicates the model explains a significant amount of variation.

6.  The assumptions of the General Linear Model must be met for the
    *p*-values to be accurate. These are: are that the relationship
    between the response and the explanatory variables is linear and
    that the residuals are normally distributed and have homogeneity of
    variance. We check these assumptions by plotting the data and the
    residuals.

7.  We use the `lm()` function to fit a linear model in R. The
    `summary()` function gives us the test statistic, *p*-values and
    *R*-squared value and the `plot()` function gives us diagnostic
    plots to check the assumptions.

8.  If the assumptions are not met we apply a non-parametric test using
    `wilcox.test` or `kruskal.test()`. These also give us a test
    statistic and a *p*-value.

9.  When reporting the results of statistical tests give the
    significance, direction and magnitude of the effect and use figures
    to demonstrate the statement.
